---
title: "STAT2005 Computer Simulation"
subtitle: "Lecture 5 — Simulation of Continuous Random Variables"
author: 
  - name: Sam Wiwatanapataphee (275287a@curtin.edu.au)
date: "2026-03-26"
date-format: "D MMM YYYY"
format:
  revealjs:
    html-math-method: mathjax
    theme: [simple, custom.scss]
    logo: figs/curtin-logo2.png
    margin: 0.05
    center: false
    slide-number: true
    incremental: true
    toc: false
    code-line-numbers: true
    code-copy: true
    scrollable: true
    title-slide: false
    embed-resources: false # true to export as single html: not compatible with chalkboard
    # scroll-view: true
    menu:
      titleSelector: 'h1, h2, h3'
      useTextContentForMissingTitles: false 
      hideMissingTitles: true
    chalkboard: true
    # Use this if you want a wider slide layout:
    # width: 1280 (default is 1050)
    # height: 720 (default is 700)
execute:
  echo: true
  warning: false
  message: false
# jupyter: python3 # cannot have both r and python
---

## Goals today

<br>

:::: {.columns}
::: {.column}
- **Part 1:** Transformation Methods (Extended)
    - Transformation of a Single Random Variable
      - Discrete Case
      - Continuous Case
        - Monotone Transformations
        - Non-Monotone Transformations
    - Inverse Transform Sampling
      - Exponential Distribution Example<br>(Closed-form Inverse CDF)
      - Quadratic CDF Example
      - Normal Distribution Example<br>(No Closed-form Inverse CDF)
      - Cubic Polynomial CDF Example
      - Limitations of Inverse Transform
:::

::: {.column}
- **Part 2:** The Acceptance-Rejection Methods
    - General Acceptance-Rejection Algorithm
    - Cubic Polynomial CDF Example<br>(Continue from Part 1)
    - Normal-Cauchy Example

<br>

- **Part 3:** Precision of Simulation Result (Revision)
    - Monte Carlo Estimation
:::

::::


# Part 1: Transformation of Random Variables


---

#### Recap: Where We Are So Far

- We use **simulation** to study random systems by repeating experiments many times
- To simulate, we need to generate **random variables**
- Random variables follow **probability distributions**

::: {.fragment style="color: blue; font-size:1.2em"}
**The Key Building Block**
:::

- Computers generate **pseudo-random numbers**
- In practice, we generate reproducible rvs: $\boxed{U \sim \text{Uniform}(0,1)}$
- Then, we used **simple rules** to generate random variables:

::: {.fragment .block .blue}
**Bernoulli(p)**

::: {.fragment .center fragment-index=5}
Generate $U \rightarrow$ If $U < p$, return 1, else return 0
:::
:::

::: {.fragment .block .green}
**Discrete Distributions:**

::: {.fragment .center fragment-index=6}
Use the precision method or the inverse CDF method
:::
:::

- **Idea:** Start with Uniform(0,1), then **transform it.**
- How do we generate **any other distribution**? Exponention? Normal? Custom distributions?

---

## Transformation of a Single Random Variable

When we apply a function to a random variable, the result is itself a new random variable. If $X$ has a known distribution and we define  
$$
Y = g(X),
$$

::: {.fragment}
then the key question is: **What is the distribution of $Y$?**
:::

::: {.fragment}
```{r}
#| fig-height: 2.5
#| fig-width: 10
#| code-fold: true
#| code-summary: Suppose $X \sim \text{Uniform}(0,1)$ and $Y = 2X, Z = X/2.$
set.seed(1234)

x <- runif(5000, 0, 1)
y <- 2 * x
z <- x / 2

par(mfrow = c(1, 3), mar = c(4,5,1,1))

hist(x, prob = TRUE, breaks = 30,
     main = "X ~ Uniform(0,1)", xlab = "x",
     col = "skyblue", border = "white",
     cex.lab = 1.4, cex.main = 1.5, cex.axis = 1.2)

hist(y, prob = TRUE, breaks = 30,
     main = "Y = 2X", xlab = "y",
     col = "tomato", border = "white",
     cex.lab = 1.4, cex.main = 1.5, cex.axis = 1.2)

hist(z, prob = TRUE, breaks = 30,
     main = "Z = X/2", xlab = "z",
     col = "seagreen2", border = "white",
     cex.lab = 1.4, cex.main = 1.5, cex.axis = 1.2)
```

:::

- If $Y = 2X \rightarrow$ values spread out (**stretched**) $\rightarrow$ density becomes lower.
- If $Y = X/2 \rightarrow$ values squeezed (**compressed**) $\rightarrow$ density becomes higher  

::: {.fragment .block .red}
🎯 **Key idea:** Probability is **conserved**, but density is **redistributed**
:::


---

### Discrete Case (Intuition)

::: {.note}
When we transform a random variable, $Y = g(X),$ multiple values of $X$ may map to the same value of $Y$.

The probability of $Y$ is obtained by **collecting contributions** from all such values:

$$
p_Y(y) = \sum_{x : g(x) = y} p_X(x).
$$

This is a “*probability bookkeeping*” rule: every value of $Y$ inherits probability from the $X$-values that produce it.

:::

<br>

::: {.fragment}

**Example:** Let $X \in \{1,2,3,4\}$ with given probabilities

$$
P(X=1)=0.1,\quad P(X=2)=0.2,
$$

$$
P(X=3)=0.3,\quad P(X=4)=0.4.
$$

Define:

$$
Y = \begin{cases}
0 & \text{if } X \text{ is odd} \\
1 & \text{if } X \text{ is even}
\end{cases}
$$

:::

::: {.fragment}

Then:

$$
Y = 0: \text{from } X = 1,3 \qquad Y = 1: \text{from } X = 2,4
$$

:::

::: {.fragment}

PMF of $Y:$

$$
P(Y=0)=P(X=1)+P(X=3)=0.1+0.3=0.4,
$$

:::

::: {.fragment}
$$
P(Y=1)=P(X=2)+P(X=4)=0.2+0.4=0.6
$$

:::

---

```{r}
#| fig-height: 3
set.seed(1234)

# Original discrete variable
X <- sample(1:4, size = 5000, replace = TRUE,
            prob = c(0.1, 0.2, 0.3, 0.4))

# Transformation
Y <- ifelse(X %% 2 == 1, 0, 1)

# Plot empirical pmf of Y
barplot(prop.table(table(Y)),
        col = c("skyblue", "tomato"),
        main = "Empirical PMF of Y = g(X)",
        ylab = "Probability")
```

---

### Continuous Case 

- For **discrete** variables: we **add probabilities.**

- For **continuous** variables: $\boxed{P(X = x) = 0.}$

::: {.fragment .center}
👉 We cannot “collect probabilities” anymore.
:::

- For continuous random variables, there are in principle two approaches: 

:::: {.columns .fragment}
::: {.column .center .block .gray width=47%}
<span style="color: gray;">**CDF‑based Method**</span>

<span style="color: gray;">derives the distribution of $Y$ by working with cumulative probabilities</span>

:::
::: {.column .center .block .blue width=47%}
<span style="color: blue;">**PDF‑based Method**</span>

determines how density changes under the transformation
:::
::::

<br>

::: {.fragment style="color: blue; font-size: 1.1em"}
**Key Idea: Density Transformation**
:::

- Instead of probabilities, we work with **density.**

- When we transform using $Y = g(X)$, we must track **how density is stretched or compressed.**

  - stretches some regions → density decreases  
  - compresses some regions → density increases  

- The adjustment is captured by a **derivative.**

---

```{mermaid}
%%| code-fold: true
%%| echo: false
%%| fig-align: center
flowchart LR
  %% Monotone transformation
  subgraph M["Monotone Y = g(X)"]
    direction TB
    x1["x₁"] -->|"g"| y1["y₁"]
    x2["x₂"] -->|"g"| y2["y₂"]
    x3["x₃"] -->|"g"| y3["y₃"]
  end

  %% Non-monotone transformation
  subgraph N["Non‑monotone Y = g(X)"]
    direction TB
    a1["x₁"] -->|"g"| b1["y₁"]
    a2["x₂"] -->|"g"| b2["y₂"]
    a3["x₃"] -->|"g"| b2
    a4["x₄"] -->|"g"| b3["y₃"]
  end

  %% Notes
  note1["Each y has exactly one preimage"]:::note
  note2["A y may have multiple preimages"]:::note

  M --- note1
  N --- note2

  %% Styling
  classDef note fill:#f5f5f5,stroke:#999,stroke-dasharray: 4 4,color:#111;

  style M fill:#f5f5f5,stroke:#999,stroke-dasharray: 4 4,color:#111;
  style N fill:#f5f5f5,stroke:#999,stroke-dasharray: 4 4,color:#111;

  %% Make all arrows black
  linkStyle default stroke:#000,stroke-width:1.5px;
```

---

### Case 1: Monotone Transformations

::: {.note}

- If $g$ is **one-to-one** and **strictly increasing or decreasing**, each value of $Y$ comes from exactly one $X.$
- Then, the PDF of $Y$ is given by the change-of-variables formula,

::: {.fragment}
$$
f_Y(y) = f_X(g^{-1}(y)) \left| \frac{d}{dy} g^{-1}(y) \right|
$$

:::
- The derivative term adjusts for **stretching/compression.**
:::

::: {.example .fragment}
**Example:** Let $X\sim \mathcal{N}(0,1)$ and define $Y=3X+2$. Determine the PDF of $Y$.

We know $Y=g(X)=3X+2$, which is strictly increasing. The inverse is

::: {.fragment}
$$
x=g^{-1}(y)=\frac{y-2}{3} \quad \text{and} \quad \frac{d}{dy}g^{-1}(y)=\frac{1}{3}.
$$

:::

::: {.fragment}
The PDF of $X$ is

$$f_X(x)=\frac{1}{\sqrt{2\pi }}e^{-x^2/2}.$$

:::

::: {.fragment}
By the change‑of‑variables formula,

$$
f_Y(y) =  f_X\! \left( \frac{y-2}{3}\right) \cdot \frac{1}{3} =\frac{1}{3}\cdot \frac{1}{\sqrt{2\pi }}\exp \! \left( -\frac{1}{2}\left( \frac{y-2}{3}\right) ^2\right).
$$

:::

::: {.fragment}
This is recognised as the PDF of $\boxed{Y\sim \mathcal{N}(2,9)}.$
:::

:::

---

```{r}
#| fig-height: 6
#| fig-align: center
#| code-fold: true
#| code-summary: Suppose $X \sim N(0,1)$ and $Y = 3X + 2$

set.seed(1234)

# Original variable
X <- rnorm(5000)

# Transformation
Y <- 3*X + 2

par(mfrow = c(2,1), mar = c(4,4,1,1))

# Theoretical X ~ N(0,1)
curve(dnorm(x, mean = 0, sd = 1), from = -10, to = 15, col = "blue", lwd = 2,
     xlab = "x", ylab = "Density", main = "X ~ N (0,1)")

# Plot simulated density
plot(density(Y), xlim =c(-10, 15),
     col = "darkred", lwd = 2,
     main = "Density of Y = 3X + 2",
     xlab = "y")

# Add theoretical curve
curve(dnorm(x, mean = 2, sd = 3),
      col = "blue", lwd = 2, add = TRUE)

legend("topright",
       legend = c("Simulated Y", "Theoretical N(2, 9)"),
       col = c("darkred", "blue"), lwd = 2, bty = "n")

```


- Because $Y=3X+2$ stretches the values of $X$, the density becomes more spread out (lower peak).

- To preserve total probability, the density is scaled by the derivative term in the change-of-variables formula:

  $$
  \left| \frac{dx}{dy} \right| = \frac{1}{3}.
  $$

---

#### Aside: Change-of-variables Formula (Intuition) {style="color: skyblue;"}



::: {.tip}
**Key Idea:** Probability is preserved under transformation, i.e.
:::

<span style="display: block; margin: 24px 0;"></span>

$$
P(a \leq Y \leq b) = P(g^{-1}(a) \leq X \leq g^{-1}(b))
$$

To understand how density changes, we look at a **small interval**:

<span style="display: block; margin: 24px 0;"></span>

$$
P(y \leq Y \leq y + dy) =
P(x \leq X \leq x + dx), \tag{1}
$$

where $\boxed{x = g^{-1}(y)}.$ From the local linear approximation (calculus), we have

$$
dx \approx \frac{dx}{dy}dy.
$$

For very small intervals, probability is approximately: **(probability ≈ density × width)**. Then,

<span style="display: block; margin: 24px 0;"></span>

$$
\begin{aligned}
P(y \leq Y \leq y + dy) &\approx f_Y(y)\;dy, \\
P(x \leq X \leq x + dx) &\approx f_X(x)\;dx.
\end{aligned} \tag{2}
$$

From (1) and (2), we have

$$
\begin{aligned}
f_Y(y) \; dy &= f_X(x) \; dx \\
f_Y(y) &= f_X(x)\frac{dx}{dy},
\end{aligned}
$$

and since densities must be positive:
$\boxed{f_Y(y) = f_X(g^{-1}(y)) \left| \dfrac{d}{dy} g^{-1}(y) \right|}$

::: {.tip}
**Takeaway:** The derivative term tells us how much the transformation stretches or compresses the space.
:::

::: notes
Linear approximation intuition
- dx is a small change in x, and dy is a small change in y
- dx/dy tells you how sensitive x is to changes in y.
- Multiplying them gives an approximation for how much x changes.
:::

---

### Case 2: Non-Monotone Transformations

::: {.note}
- If $g$ is *not* one-to-one, multiple values of $X$ map to the same $Y.$

- Then we must **add all contributions**:

::: {.fragment}
$$
f_Y(y) = \sum_{x : g(x) = y}
f_X(x)\left|\frac{dx}{dy}\right|
$$

:::
:::

<br>

::: {.example .fragment}
**Example:** Let $X \sim \mathcal{N}(0,1)$ and define $Y = X^2.$ Find the PDF of $Y$.

- For a given $y$, two values map to it:

::: {.fragment}
$$
x = \sqrt{y}, \quad x = -\sqrt{y}
$$

:::

- Both contribute to the density
- Using the non-monotone trasformation rule, we have:

::: {.fragment}
$$
f_Y(y) = \frac{1}{\sqrt{2\pi}}e^{-y/2}\cdot\frac{1}{\sqrt{y}}.
$$

:::

- This is recognised as the PDF of a Chi-squared distribution with 1 degree of freedom, i.e. $\boxed{Y \sim \chi^2_1}$

:::

::: aside
See **Transformation of Random Variables** course note for a full derivation of $f_Y(y).$
:::

---

```{r}
#| fig-height: 6
#| fig-align: center
#| code-fold: true
#| code-summary: Suppose X ~ N(0,1) and Y = X^2
set.seed(1234)

# Original variable
X <- rnorm(5000)

# Transformation
Y <- X^2

op <- par(mfrow = c(2,1), mar = c(4,4,2,1))

# Panel 1: X ~ N(0,1)
curve(dnorm(x, mean = 0, sd = 1),
      from = -4, to = 4,
      col = "blue", lwd = 2,
      xlab = "x", ylab = "Density",
      main = "Density of X ~ N(0,1)")

# Panel 2: Y = X^2
hist(Y, breaks = 40, freq = FALSE,
     col = "lightgray", border = "white",
     xlim = c(0, 6), ylim = c(0, 1.1),
     main = "Density of Y = X^2",
     xlab = "y", ylab = "Density")

curve(dchisq(x, df = 1),
      from = 1e-4, to = 6,
      col = "darkred", lwd = 2, add = TRUE)

legend("topright",
       legend = c("Simulated Y", expression(chi[1]^2)),
       col = c("lightgray", "darkred"),
       lwd = c(10, 2), bty = "n")
```

- $X \sim N(0,1)$ distribution spreads across both negative and positive values.
- $Y = X^2$ distribution is right-skewed (non-negative).

---

#### Why Transformation Matters to Simulation


So far:

::: {.fragment}
<div class="flowchart" aria-label="Transformation">
  <div class="node" style="background: rgba(0,90,180,0.08);">We start with $X$</div>
  <div class="arrow" aria-hidden="true">→</div>

  <div class="node" style="background: rgba(0, 120, 60, 0.08);">Apply transformation $Y = g(X)$</div>
  <div class="arrow" aria-hidden="true">→</div>

  <div class="node" style="background: rgba(170, 160, 0, 0.10);">Find distribution of $Y$</div>
</div>

:::

<br>

::: {.fragment}
Now we reverse the idea:
:::

::: {.fragment}
<div class="flowchart" aria-label="Transformation">
  <div class="node" style="background: rgba(0,90,180,0.08);">Start with $U \sim \text{Uniform}(0,1)$</div>
  <div class="arrow" aria-hidden="true">→</div>


  <div class="node" style="background: rgba(170, 160, 0, 0.10);">Choose $g$ to create a desired distribution</div>
</div>

:::

<br>

::: {.block .yellow .fragment}

👉 **Goal:**

$$
X = g(U)
$$

so that $X$ has the distribution we want

:::

::: {.fragment}
How do we find such a transformation $g$?
:::

::: {.center .fragment style="margin-top: 5px; font-size: 1.2em;"}
Use the **CDF**
:::

::: {.fragment}
This leads to
:::

::: {.center .fragment style="margin-top: 5px; font-size: 1.2em; color: blue;"}
**Inverse Transform Sampling**
:::

---

## Inverse Transform Sampling

::: {.fragment}

Every random variable has a CDF:

$$
F_X(x) = P(X \le x)
$$

:::

::: {.fragment}
The CDF maps values of $X$ → probabilities in (0,1)
:::

::: {.fragment}

> 👉 This suggests a connection to **Uniform(0,1)**

:::

<span style="display: block; margin: 24px 0;"></span>

### Probability Integral Transform {.fragment}

::: {.fragment}

If $X$ has CDF $F_X$, then:

$$
U = F_X(X) \sim \text{Uniform}(0,1)
$$

:::

::: {.fragment}

> 👉 The CDF transforms **any distribution → Uniform(0,1)**

:::

::: {.fragment}

then reversing gives:

$$
\boxed{X = F_X^{-1}(U), \quad U \sim \text{Uniform}(0,1)}
$$

which is known as the **sampling rule**.
:::

::: {.note .fragment}

To simulate $X \sim F_X$:

::: {.fragment}
<div class="flowchart" aria-label="Transformation">
  <div class="node" style="background: rgba(0,90,180,0.08);">Generate $U \sim \text{Uniform}(0,1)$</div>
  <div class="arrow" aria-hidden="true">→</div>

  <div class="node" style="background: rgba(170, 160, 0, 0.10);">Transform: $X = F_X^{-1}(U)$</div>
</div>

:::
:::

---


### Exponential Distribution Example (Closed-form Inverse CDF)

::: {.example}

Let $X \sim \text{Exp}(\lambda)$ with the following CDF:

$$
F(x) = 1 - e^{-\lambda x}
$$

Setting $U = F(X)$ and solving for $X$ gives the inverse term:


$$
X = -\frac{1}{\lambda} \log(1 - U), \quad U \sim \text{Uniform}(0,1).
$$


:::

::: {.fragment}

```{r}
#| fig-height: 4
#| fig-align: center
#| code-fold: true
#| code-summary: X ~ Exp($\lambda=2$)
set.seed(1234)

u <- runif(1000)
lambda <- 2

x <- -log(1 - u) / lambda

hist(x, prob = TRUE, col = "skyblue",
     main = "Simulated Exponential Samples")

curve(dexp(x, rate = lambda),
      col = "red", lwd = 2, add = TRUE)

legend("topright",
       legend = c("Simulated Exp", "Theoretical Exp"),
       col = c("skyblue", "red"), lwd = c(10,2), bty = "n")
```

:::


---

### Quadratic CDF Example

::: {.example}
Consider the **electrical current distribution** model where the pdf of $X$ is given by


$$
f(x) = 1.25 - 0.25x, \quad 2 \leq x \leq 4.
$$


Suppose a simulation of $X$ is required as part of some larger system analysis.

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
<div class="node" style="background: rgba(135, 0, 180, 0.08); width: 13em;">Step 0: Check for valid PDF</div>

:::

::: {.fragment}
$$
\int_2^4 (1.25 - 0.25x)\,dx = 1 \quad \rightarrow \quad \text{valid density.}
$$

:::


::: {.fragment}
<div class="node" style="background: rgba(0,90,180,0.08); width: 15em;">Step 1: Determine the CDF $F(x)$</div>

:::

::: {.fragment}
$$
\begin{aligned}
F(x)
&= \int_2^x (1.25 - 0.25y)\,dy \\
&= \left[1.25y - 0.125y^2 \right]_2^x \\
F(x) &= 1.25x - 0.125x^2 - 2
\end{aligned}
$$

:::

::: {.fragment}
<div class="node" style="background: rgba(0, 180, 57, 0.08); width: 23em;">Step 2: Determine the inverse CDF $X=F^{-1}_X(U)$</div>

:::

::: {.fragment}
$$
\begin{aligned}
u &= F(x) = 1.25 x - 0.125x^2 -2 \\
0 &= x^2 - 10x + 16 + 8u \\
x &= 5 \pm \sqrt{9-8u}
\end{aligned}
$$

:::

- Since the support is $2 \leq x \leq 4,$ the only root satisfies this is $\boxed{x = 5 - \sqrt{9 - 8u}}.$

---

- $\boxed{X = 5 - \sqrt{9 - 8U}}.$
- This transformation maps Uniform(0,1) values into the interval [2,4] with the correct shape.

::: {.fragment}

```{r}
#| fig-height: 4
set.seed(1234)
u <- runif(10000)
x <- 5 - sqrt(9 - 8 * u)

hist(x, prob = TRUE, col = "skyblue",
     main = "Simulated Distribution")

curve(1.25 - 0.25*x, add = TRUE, col = "red", lwd = 2)
```

:::

---

### Normal Distribution Example (No Closed-form Inverse CDF)

::: {.example}
Recall the PDF of the **standard normal distribution** $X \sim N(0,1):$

<span style="display: block; margin: 24px 0;"></span>

$$
f(x)=\frac{1}{\sqrt{2\pi}}\exp{\left(-\frac{x^2}{2}\right)}, \quad x \in \mathbb{R}.
$$

<span style="display: block; margin: 24px 0;"></span>

The CDF of $X$ is $\Phi(x):$

<span style="display: block; margin: 24px 0;"></span>

$$
\Phi(x) = \frac{1}{\sqrt{2\pi}}\int_{-\infty}^x\exp{\left(-\frac{y^2}{2}\right)}\,dy,
$$

<span style="display: block; margin: 24px 0;"></span>

has <u>no closed-form expression</u>.

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}


<span style="display: block; margin: 24px 0;"></span>

- The CDF has no closed-form expression, so neither does its inverse.
  - We must compute it numerically.
- This is the reason why we use software or table to calculate Normal probabilities.

:::

:::

---

### Cubic Polynomial CDF Example

::: {.example}
$F(x)$ is a **polynomial function.**

<span style="display: block; margin: 24px 0;"></span>

The measurement error $X$ (in mV) of a particular volt-meter has the following distribution:

<span style="display: block; margin: 24px 0;"></span>

$$
f(x) = 
\begin{cases}
\dfrac{4-x^2}{9}, & \quad -1 \leq x \leq 2 \\
0, & \text{otherwise.}
\end{cases}
$$

::: {.fragment}

The CDF is

$$
F(x) = \frac{-x^3+12x+11}{27}, \quad -1 \leq x \leq 2,
$$

<span style="display: block; margin: 24px 0;"></span>

which is a cubic polynomial function.


:::

<span style="display: block; margin: 24px 0;"></span>

- It is not easy to solve for $x = F^{-1}(u).$

- Numerical methods (e.d. Newton–Raphson) should be used to solve this kind of problem
:::

---

#### When Does Inverse Transform Work?

<br>

::: {.center .fragment}
Even if we *can* solve for inverse CDF, it may be:
:::

::: {.center}
- slow
- unstable
- impractical in simulation
:::

<span style="display: block; margin: 24px 0;"></span>



| Case | Difficulty |
|------|------------|
| Exponential | ✔ Easy (closed-form inverse) |
| Electrical example | ✔ Moderate (quadratic) |
| Normal | ❌ No closed-form inverse |
| Polynomial CDF | ⚠️ Requires numerical methods |

: {.striped .fragment}

<br>

::: {.callout-note .fragment style="font-size:1.4em;"}
👉 Not all distributions are equally tractable.
:::


---

### Limitations of Inverse Transform

<span style="display: block; margin: 24px 0;"></span>

::: {.block .yellow .fragment}
**No closed-form inverse**

Many distributions (e.g. Normal) do not have an explicit inverse CDF.
:::


<span style="display: block; margin: 24px 0;"></span>

::: {.block .blue .fragment}
**Inversion can be difficult**

Even when $F(x)$ is known, solving $u = F(x)$ may require numerical methods.
:::

<span style="display: block; margin: 24px 0;"></span>

::: {.block .purple .fragment}
**Computational cost**

Numerical inversion can be slow or impractical in large simulations.
:::


<span style="display: block; margin: 24px 0;"></span>

::: {.block .green .fragment}
👉 **We need more general methods** that does not reqire $F^{-1}$ and work for complex distributions.

- <span style="color: blue;">**Acceptance–Rejection Sampling**</span>
- Specialised transformations such as **Box–Muller** for Normal simulation
:::


# Part 2: The Acceptance-Rejection Methods

## Acceptance-Rejection Sampling

:::: {.columns}

::: {.column width=55%}

::: {.example}
**Example:** We want to simulate

$$
f(x) = 3x^2, \quad 0 \le x \le 1,
$$

which has increasing density with more weight near 1.

How do we simulate this?

:::

<span style="display: block; margin: 24px 0;"></span>

#### First, check for valid PDF.

$$
\int_0^1 3x^2 \, dx = 1 \rightarrow \text{ valid PDF.}
$$


#### Why Not Inverse Transform? {.fragment}

- We would need $X = F^{-1}(U).$
- Finding $F(x)$ is easy, but inverting it is not simple.
- 👉 Let’s try a different idea.

:::
::: {.column width=43%}

```{r}
#| fig-height: 6
x <- seq(0, 1, length.out=100)
par(mar=c(4,5,1,0))
curve(3 * x^2, type = "l",
      lwd = 3, ylab = "f(x)",
      cex.axis=2, cex.lab=2.5)
```

:::
::::



#### Key Idea {.fragment}

- Instead of transforming, we **sample points in a region.**

::: {.fragment}
<div class="flowchart" aria-label="Transformation">
  <div class="node" style="background: rgba(0,90,180,0.08);">Draw a random point $(X, Y)$</div>
  <div class="arrow" aria-hidden="true">→</div>

  <div class="node" style="background: rgba(170, 160, 0, 0.10);">Accept it if it lies under the curve</div>
</div>

:::

- 👉 This turns sampling into a **geometry problem.**

---

#### Visualisation

:::: {.columns}

::: {.column}

Consider the rectangle:

$$
0 \le x \le 1, \quad 0 \le y \le 3.
$$

::: {.fragment fragment-index=2}

The curve $y = 3x^2$ sits inside this box.

:::


::: {.fragment fragment-index=3}
👉 Accept points under the curve  
👉 Reject points above it

:::

<br>


#### Acceptance–Rejection Algorithm {.fragment fragment-index=4}

::: {.note .fragment}
1. Generate:
   - $X \sim \text{Uniform}(0,1)$
   - $Y \sim \text{Uniform}(0, 3)$

2. If:
  $$
  Y \le 3X^2
  $$

::: {.fragment}
→ Accept $X$
:::

3. Otherwise → Reject and repeat

:::
:::

::: {.column}

::: {.fragment fragment-index=1}

```{r}
#| fig-height: 5.5
#| echo: false
x <- seq(0, 1, length.out=100)
par(mar=c(4,5,1,0))
curve(3 * x^2, type = "l",
      lwd = 3, ylab = "f(x)",
      cex.axis=2, cex.lab=2.5)
abline(v=0, lwd = 3, col = "blue")
abline(v=1, lwd = 3, col = "blue")
abline(h=0, lwd = 3, col = "blue")
abline(h=3, lwd = 3, col = "blue")
```

:::

#### Why Does This Work? {.fragment}

- Points are uniformly distributed in the rectangle.
- Probability of acceptance:

  > proportional to area under the curve.

::: {.fragment}
👉 Accepted $X$ values follow the target distribution.
:::

:::


::::

---

```{r}
# Visual points
x <- runif(2000)
y <- runif(2000, 0, 3)

plot(x, y, pch = 16, col = ifelse(y <= 3*x^2, "blue", "red"))
curve(3*x^2, add = TRUE, col = "black", lwd=3)
```

---

```{r}
#| output: false
set.seed(1234)

n <- 10000

accepted <- numeric(0) # initialise numeric variables/vectors

while(length(accepted) < n){ # keep sampling until we have n accepted values
  x <- runif(1) # proposal: X ~ Uniform(0,1)
  y <- runif(1, 0, 3) # auxiliary variable: Y ~ Uniform(0, 3)
  
  if(y <= 3 * x^2){ # accept if point lies under the curve f(x)
    accepted <- c(accepted, x) # keep accepted x
  } # otherwise reject and try again
}

hist(accepted, prob = TRUE, col = "skyblue", main = "Acceptance-Rejection Sampling")
lines(density(accepted), col = "blue", lwd = 2) # Kernel Density Estimate (KDE)
curve(3 * x^2, col = "red", lwd = 2, add = TRUE) # True Density

legend("topright", legend = c("Histogram", "KDE", "True density"),
       col = c("skyblue", "blue", "red"),
       lwd = c(NA, 2, 2), pch = c(15, NA, NA), bty = "n")
```

---


```{r}
#| echo: false
set.seed(1234)

n <- 10000

accepted <- numeric(0) # initialise numeric variables/vectors

while(length(accepted) < n){ # keep sampling until we have n accepted values
  x <- runif(1) # proposal: X ~ Uniform(0,1)
  y <- runif(1, 0, 3) # auxiliary variable: Y ~ Uniform(0, 3)
  
  if(y <= 3 * x^2){ # accept if point lies under the curve f(x)
    accepted <- c(accepted, x) # keep accepted x
  } # otherwise reject and try again
}

hist(accepted, prob = TRUE, col = "skyblue", main = "Acceptance-Rejection Sampling")
lines(density(accepted), col = "blue", lwd = 2) # Kernel Density Estimate (KDE)
curve(3 * x^2, col = "red", lwd = 2, add = TRUE) # True Density

legend("topleft", legend = c("Histogram", "KDE", "True density"),
       col = c("skyblue", "blue", "red"),
       lwd = c(NA, 2, 2), pch = c(15, NA, NA), bty = "n")
```


- The histogram matches the increasing shape of the target density, so larger values near 1 are more likely. 
- Small differences between bars and the curve are due to randomness and binning.

---

#### From Rectangles to General Envelopes

In the first example, we sampled uniformly from a rectangle and accepted points under the curve.

That works well when:

- the support is simple
- the density is bounded
- a rectangle is easy to use

::: {.fragment}
But many target densities are more complicated. Therefore, we need a more flexible idea.
:::

<br>

#### Key Idea {.fragment}

::: {.note .fragment}

::: {.fragment}
Suppose we want to simulate from a target density $f(x)$.
Instead of sampling from a rectangle, we choose:
:::

- an easier density $g(x)$
- a constant $M > 0$

::: {.fragment}
such that

$$
\boxed{f(x) \le M g(x) \quad \forall x}
$$
:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
Then $Mg(x)$ acts as an **envelope** over $f(x).$
:::

::: {.fragment}
👉 The closer $Mg(x)$ is to $f(x)$, the more efficient the method.
:::

:::

---

### General Acceptance-Rejection Algorithm

::: {.note}

To simulate $X \sim f(x):$

::: {.fragment}
<div class="flowchart" aria-label="Transformation">
  <div class="node" style="background: rgba(0,90,180,0.08); text-align: center;">Generate a candidate<br>$Y \sim g(x)$</div>
  <div class="arrow" aria-hidden="true">→</div>

  <div class="node" style="background: rgba(135, 0, 180, 0.08); text-align: center;">Generate<br>$U \sim \text{Uniform}(0,1)$</div>
  <div class="arrow" aria-hidden="true">→</div>

  <div class="node" style="background: rgba(9, 180, 0, 0.08); text-align: center;">Accept $Y$ if<br>$U \le \dfrac{f(Y)}{M g(Y)}$</div>
  <div class="arrow" aria-hidden="true">→</div>

  <div class="node" style="background: rgba(170, 0, 0, 0.1); text-align: center;">Otherwise<br>reject $Y$ and repeat</div>
</div>

:::

- If the candidate $Y$ is accepted, then we set $X = Y$.
- The accepted values follow the target density $f(x)$.

:::

<span style="display: block; margin: 24px 0;"></span>

### Efficiency of the Method {.fragment}

::: {.fragment}
The probability of acceptance is

$$
\frac{1}{M}.
$$

:::

- smaller $M$  $\rightarrow$ higher acceptance rate
- larger $M$  $\rightarrow$ more rejections

### Choosing a Good Proposal Density {.fragment}

:::: {.columns}

::: {.column}
- have the same support as $f(x)$
- be easy to sample from
- make $M$ as small as possible
:::
::: {.column}

- Uniform proposal on a bounded interval
- Exponential proposal for right-skewed targets
- Normal proposal for bell-shaped targets
:::
::::

::: notes
Example on Efficient of the Method:
- f(x) = 2x, 0 <= x <= 1.
- f(x):straight line with slope 2 and y-intercept at 0.
- easy g(x) = 1 & good M = 2
- Mg(x) = 2: horizontal line at y=2.
- accept area = reject area = triangle with probability = 1/M = 0.5.
- with bad M, says M = 10
- Mg(x) = 10: horizontal line at y=10.
- accept area = 1/2 * 1 * 2 = 1
- reject area = 1/2 * (10 + 8) * 1 = 9
- accept prob = 1/M = 1/10
:::

---

### Cubic Polynomial CDF Example (From Part 1)

:::: {.columns}

::: {.column}
::: {.example}
The measurement error $X$ (in mV) of a particular volt-meter has the following distribution:

<span style="display: block; margin: 24px 0;"></span>

$$
f(x) = 
\begin{cases}
\dfrac{4-x^2}{9}, & \quad -1 \leq x \leq 2 \\
0, & \text{otherwise.}
\end{cases}
$$

Simulate $f(x)$ using acceptance-rejection method.
:::

:::

::: {.column}

```{r}
#| fig-height: 4
#| echo: false
x <- seq(-1, 2, length.out=100)
par(mar=c(4,5,0,0))
curve((4 - x^2)/9, xlim = c(-1,2), type="l", lwd=3, ylab="f(x)",
      cex.axis=1.5, cex.lab=2)
```

:::

::::

::: {.fragment}
<div class="node" style="background: rgba(0,90,180,0.08); width: 17em;">Step 0: Check whether $f(x)$ is valid</div>

$$
\int_{-1}^2 \frac{4-x^2}{9} \,dx = 1 \rightarrow \text{ valid PDF.}
$$

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
<div class="node" style="background: rgba(0, 180, 48, 0.08); width: 20em;">Step 1: Determine $Mg(x); f(x) \leq Mg(x)$ </div>

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
Because the support is bounded, the best first proposal is:

$$
g(x) = \text{Uniform}(-1, 2) = \frac{1}{3}, \quad -1 \leq x \leq 2.
$$

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
Since $g(x)=1/3$, this becomes

$$
\frac{4-x^2}{3} \leq M
$$

:::

---


Now on [-1, 2], the maximum of LHS occurs at $x=0,$ 

$$
\max_{-1\leq x \leq 2} \frac{4-x^2}{3} = \frac{4}{3}
$$


::: {.fragment}
Hence, we can choose

$$
M = \frac{4}{3} \quad \text{ with an acceptance rate of } \quad \frac{1}{M} = \frac{3}{4} = 75\%
$$

:::

::: {.fragment}
This gives

$$
Mg(x) = \frac{4}{3} \cdot \frac{1}{3} = \frac{4}{9}
$$

:::

- So the envelope is just a horizontal line at height 4/9.

::: {.fragment}
```{r}
#| fig-height: 3.2
#| fig-width: 6
#| fig-align: center
#| echo: false
x <- seq(-1, 2, length.out=100)
par(mar=c(4,5,0,0))
curve((4 - x^2)/9, xlim = c(-1,2), type="l", lwd=3, ylab="f(x)")
abline(h=4/9, col="blue", lwd=3, lty = 2)

legend("topright",
       legend = c("f(x)", "Mg(x) = 4/9"),
       col = c("black", "blue"),
       lwd = c(3, 2), lty = c(1, 2), bty = "n")
```

:::

---


<div class="node" style="background: rgba(0, 180, 48, 0.08); width: 18em;">Step 2: Determine the acceptance rule </div>

The general rule is

$$
U \leq \frac{f(Y)}{Mg(Y)}
$$

where $Y\sim g$ and $U \sim \text{Uniform}(0,1).$

::: {.fragment}
Substitute in, gives

$$
\frac{f(Y)}{Mg(Y)} = \frac{4-y^2}{4}
$$

:::

::: {.fragment}
So accept $Y$ if

$$
U \leq \frac{4-y^2}{4}.
$$

:::

::: {.fragment}

```{r}
set.seed(1234)
n <- 10000
x <- numeric(0) # store accepted samples

while(length(x) < n){ # keep sampling until we have n accepted values
  y <- runif(n, min = -1, max = 2)   # proposal sample Y ~ g(x) ~ Uniform(-1, 2)
  u <- runif(n)                      # U ~ Uniform(0,1) for acceptance test
  accept <- u <= (4 - y^2)/4         # acceptance rule (returns TRUE/FALSE)
  x <- c(x, y[accept])               # keep accepted y
}

x <- x[1:n] # trim to exactly n samples
```

:::

---

```{r}
#| fig-align: center
hist(x, prob = TRUE, breaks = 40,
     col = "lightgray", border = "white",
     main = "Acceptance-Rejection Sampling",
     xlab = "x")

curve((4 - x^2)/9, from = -1, to = 2,
      add = TRUE, col = "red", lwd = 2)
```

---

### Normal-Cauchy Example

::: {.example}
Generate the standard normal distribution using the standard Cauchy distribution

Recall the PDF of $X \sim N(0, 1):$

$$
f(x)=\frac{1}{\sqrt{2\pi}}\exp{\left(-\frac{x^2}{2}\right)}, \quad x \in \mathbb{R}.
$$

The PDF of $X \sim \text{Cauchy}:$

$$
g(x) = \frac{1}{\pi(1+x^2)}, \quad x \in \mathbb{R}.
$$

:::

Let's compare the distributions:

```{r}
#| code-fold: true
#| fig-align: center
#| fig-height: 3
x <- seq(-5, 5, length = 200)
dc <- dcauchy(x, location = 0, scale = 1)
dn <- dnorm(x, mean = 0, sd = 1)

par(mfrow=c(1,2), mar=c(4,5,1,1))
plot(x, dn, type="l", ylab="Density", cex.lab=1.5, cex.axis=1.5)
lines(x, dc, col="blue", lty=2)
legend("topright", col=c("black", "blue"), lty=c(1,2), legend =c("Normal", "Cauchy"))
plot(x, dn/dc, type="l", ylab="f(x) / g(x)", cex.lab=1.5, cex.axis=1.5)
```


- From RHS, the standard cauchy distribution is wider but shorter than the standard normal distribution.
- From LHS, the $f(x)/g(x)$ ratio is maximised around 1.5.


---

<div class="node" style="background: rgba(0, 180, 48, 0.08); width: 7em;">Step 1: Find $M$ </div>

$$
\frac{f(x)}{g(x)} = \frac{1}{\sqrt{2\pi}}\exp{\left(-\frac{x^2}{2}\right)} \cdot \pi(1+x^2) = \frac{\pi}{\sqrt{2\pi}}(1+x^2)\exp{\left(-\frac{x^2}{2}\right)}
$$

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
$$
\max_{-\infty < x < \infty}{(1+x^2)\exp{\left(-\frac{x^2}{2}\right)}}
$$

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
Differentiate and set it equal to 0:

<span style="display: block; margin: 24px 0;"></span>

$$
\begin{aligned}
\frac{d}{dx} \left(\frac{f(x)}{g(x)}\right) &= 2x\exp{\left(-\frac{x^2}{2}\right)} + (1+x^2)(-x)\exp{\left(-\frac{x^2}{2}\right)} \\
0 &= \exp{\left(-\frac{x^2}{2}\right)} x (1-x^2) 
\end{aligned}
$$

yields three critical points: $x = 0, x = \pm 1.$

:::

<span style="display: block; margin: 24px 0;"></span>

- From the ratio plot, we would expect the maximum to occur at $x = \pm 1.$

- If this is not obvious, use the second derivative test to verify.

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
Substitute $x= \pm 1$ into $f(x)/g(x)$ yields

$$
M=2e^{-1/2}\sqrt{\frac{\pi}{2}}\approx 1.5203 \rightarrow \text{ with the acceptance probability of } 0.6577.
$$

:::

---

<div class="node" style="background: rgba(0, 180, 48, 0.08); width: 18em;">Step 2: Determine the acceptance rule </div>

The general rule is

$$
U \leq \frac{f(Y)}{Mg(Y)}
$$

where $Y\sim g \sim \text{Cauchy}$ and $U \sim \text{Uniform}(0,1).$

::: {.fragment}
Substitute in, gives the acceptance rule

$$
U \leq  \frac{f(Y)}{Mg(Y)} = \frac{\sqrt{e}}{2} (1+y^2) \exp{\left(-\frac{y^2}{2}\right)}
$$

:::

::: {.fragment}
```{r}
set.seed(1234)
n <- 5000              # number of samples needed
M <- 2 * exp(-1/2) * sqrt(pi/2)   # exact M ≈ 1.5203
samples <- numeric(0)

while(length(samples) < n){
  y <- rcauchy(n) # propose from Cauchy
  u <- runif(n)   # uniform for acceptance
  f <- dnorm(y)   # target density
  g <- dcauchy(y) # proposal density
  
  accept <- u <= f / (M * g) # acceptance rule
  samples <- c(samples, y[accept])  # keep accepted values
}

samples <- samples[1:n] # trim to exactly n samples
```

:::

---

```{r}
#| fig-height: 3
#| fig-align: center
hist(samples, prob = TRUE, breaks = 40,
     col = "lightgray",
     main = "Normal via Acceptance-Rejection (Cauchy proposal)",
     xlab = "x")

curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
```

```{r}
set.seed(1234)
y <- rcauchy(5000); u <- runif(5000)
M <- 2 * exp(-1/2) * sqrt(pi/2)
accept <- u <= dnorm(y) / (M * dcauchy(y))
cat("The empirical acceptance rate is", mean(accept))
```

---

# Part 3: Precision of Simulation Result



::: {.fragment .center style="font-size: 1.1em;"}
Up until this point, we learned how to generate random variables using

- known distributions
- inverse transform
- acceptance-rejection

:::

::: {.fragment .center style="font-size: 1.1em;"}
We can now simulate almost any distribution.
:::

<br>

::: {.fragment .center style="font-size: 1.5em; color: blue;"}
How accurate are our simulation results?
:::

<br>

::: {.fragment .center style="font-size: 1.1em;"}
- If we simulate 1,000 samples... is that enough?
- What about 10,000?
- How close are we to the true value?
:::

<br>

::: {.fragment .center style="font-size: 1.1em;"}
When we use these simulations to estimate quantities,

this is called <span style="color:blue;">**Monte Carlo estimation**</span>.
:::

---

## Monte Carlo Estimation


Suppose we want to estimate:

$$
\mu = \mathbb{E}[X]
$$

We simulate i.i.d random variables:

$$
X_1, X_2, \dots, X_n
$$

and compute:

$$
\bar{X}_n = \frac{1}{n}\sum_{i=1}^n X_i,
$$

where $n$ is the number of simulated samples.

<br>

#### Key Result (CLT)


The CLT tells us about the distribution of $\bar{X}_n$:

$$
\bar{X}_n \approx \mathcal{N}\left(\mu, \frac{\sigma^2}{n}\right)
$$


based on the same $n$ samples.

<br>

::: {.tip}
The CLT tells us that the error of Monte Carlo estimate is approximately Normally distributed for large $n$, which allows us to construct **confidence intervals** and *quantify simulation precision.*
:::

---

#### Monte Carlo Standard Error

The **standard error** measures simulation uncertainty:

$$
\text{SE}(\bar{X}_n) = \frac{\sigma}{\sqrt{n}}
$$

In practice:

$$
\widehat{\text{SE}} = \frac{s}{\sqrt{n}}
$$

where $s$ is the sample standard deviation.

**Interpretation:** Larger $n$ → smaller error, in which error decreases at rate $1/\sqrt{n}.$

<br>

```{r}
set.seed(1234)

n <- 1000
x <- rexp(n, rate = 2)
x2 <- rexp(4*n, rate = 2)

mean(x); mean(x2)
sd(x)/sqrt(n); sd(x2)/sqrt(4*n)
```

---

#### Confidence Interval for the Mean

:::: {.columns}

::: {.column .center}

CI for sample mean is

<br>

$$
\bar{X}_n \pm z_{\alpha/2} \frac{s}{\sqrt{n}}
$$

:::

::: {.column .center}

| Confidence Level<br>$(1-\alpha)100\%$ | $z_{\alpha/2}$ |
|:-------:|:--------:|
| 90% | $1.645$ |
| 95% | $1.960$ |
| 99% | $1.645$ |

: {.striped}

::::


```{r}
#| code-fold: true
#| fig-height: 4
#| fig-align: center
set.seed(1234)

n <- 1000
B <- 1000

means <- replicate(B, mean(rexp(n, rate=2)))

hist(means, prob=TRUE, col="lightgray",
     main="Sampling Distribution of the Mean")

mean_est <- mean(means)
se <- sd(means)

abline(v = mean_est, col="red", lwd=2)
abline(v = mean_est + c(-1,1)*1.96*se,
       col="blue", lty=2, lwd=2)
```