---
title: "STAT2005 Computer Simulation"
subtitle: "Lecture 6 — Multivariate Simulation"
author: 
  - name: Sam Wiwatanapataphee (275287a@curtin.edu.au)
date: "2026-04-02"
date-format: "D MMM YYYY"
format:
  revealjs:
    html-math-method: mathjax
    theme: [simple, custom.scss]
    logo: figs/curtin-logo2.png
    margin: 0.05
    center: false
    slide-number: true
    incremental: true
    toc: false
    code-line-numbers: true
    code-copy: true
    scrollable: true
    title-slide: false
    embed-resources: false # true to export as single html: not compatible with chalkboard
    # scroll-view: true
    menu:
      titleSelector: 'h1, h2, h3'
      useTextContentForMissingTitles: false 
      hideMissingTitles: true
    chalkboard: true
    # Use this if you want a wider slide layout:
    # width: 1280 (default is 1050)
    # height: 720 (default is 700)
execute:
  echo: true
  warning: false
  message: false
# jupyter: python3 # cannot have both r and python
---

## Goals today

<br>


- **Part 1:** Transformation of Multiple RVs
    - Transformations of Two Random Variables
    - The Jacobian Formula

<br>

- **Part 2:** Linear Transformations
    - Transformations of Random Vectors
    - Multivariate Normal Distribution
    - Cholesky

<br>

- **Part 3:** Copulas
    - Gaussian Copula
    - Case Study: Portfolio risk simulation


# Part 1: Transformation of Multiple RVs

## Why Multivariate Simulation

<span style="display: block; margin: 24px 0;"></span>

::: {.center}
In real-world problems, we rarely simulate a single variable in isolation.
:::

::: {.fragment .block .green .center}
**Financial models** → multiple asset returns
:::

::: {.fragment .example}
**Example:** Suppose you are managing a portfolio with three assets: Apple, Microsoft, and Tesla. Tech stocks usually move together, so we must simulate them jointly.
:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment .block .red .center}
**Reliability studies** → lifetimes of related components  
:::

::: {.fragment .example}
**Example:** Consider a car engine system with multiple components: battery, alternator, and starter motor. Each component has its own lifetime, but their lifetimes tend to fail together or degrade together (dependent).
:::

<span style="display: block; margin: 24px 0;"></span>


::: {.fragment .block .yellow .center}
**Service systems** → arrivals, service times, behaviour
:::

::: {.fragment .example}
**Example:** Consider a busy coffee shop during the morning rush. We need to simulate three *interacting* components: arrivals, service times, and customer behaviours.
:::

<span style="display: block; margin: 24px 0;"></span>

#### Key Challenge {.fragment .center}

::: {.fragment .center}
Simulating each variable separately is **not enough**.

We must also model **dependence structure** and **joint behaviour**
:::

---

#### How Do We Do This?

::: {.fragment .center}
We build from simple to powerful tools:
:::

::: {.center}
- Transformations of two random variables  
- Linear transformations → correlation  
- Multivariate normal distribution  
- Cholesky decomposition → simulation  
- Copulas → flexible dependence modelling  
:::


## Transformations of Two Random Variables

::: {.note}
Suppose $(X, Y)$ has a known joint density $f_{X,Y}(x,y)$, and we define a new pair:
<span style="display: block; margin: 24px 0;"></span>

$$
U = g_1(X, Y), \qquad V = g_2(X, Y).
$$

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
Our goal is to find the joint density $f_{U,V}(u,v).$
:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
If the transformation is one‑to‑one and differentiable, we can invert it:

<span style="display: block; margin: 24px 0;"></span>
$$
x = h_1(u,v), \qquad y = h_2(u,v),
$$

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
and use the **Jacobian determinant** to adjust for how the transformation stretches or compresses area.
:::
<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
If the transformation is smooth and invertible, then

<span style="display: block; margin: 24px 0;"></span>

$$
\boxed{f_{U,V}(u,v)
= f_{X,Y}(h_1(u,v),\, h_2(u,v))
\left| \det J \right|}
$$

:::

::: {.fragment}
where

$$
J = 
\begin{pmatrix}
\dfrac{\partial x}{\partial u} & \dfrac{\partial x}{\partial v} \\
\dfrac{\partial y}{\partial u} & \dfrac{\partial y}{\partial v}
\end{pmatrix}
$$

is the **Jacobian matrix** of the inverse transformation.
:::

::: {.fragment}
The $|\det J|$ measures how the transformation changes area, ensuring that total probability remains 1.
:::


:::


---

### Example 1: Sums and Differences

Let $(X,Y)$ have joint density $f_{X,Y}(x,y)$. Define


$$
U = X + Y, \qquad V = X - Y.
$$


This transformation is linear and invertible, and the Jacobian is constant.  

::: {.fragment}
**Step 1: Invert the transformation**

Solve for $X, Y$ in terms of $U, V$:
:::

::: {.fragment}
$$
X = \dfrac{U + V}{2}, \qquad Y = \dfrac{U - V}{2}.
$$

:::

::: {.fragment}
So

$$
x = h_1(u,v) = \dfrac{u+v}{2}, \quad
y = h_2(u,v) = \dfrac{u-v}{2}.
$$

:::

::: {.fragment}
**Step 2: Compute the Jacobian**
:::

::: {.fragment}
$$
J =
\begin{pmatrix}
\dfrac{\partial x}{\partial u} & \dfrac{\partial x}{\partial v} \\
\dfrac{\partial y}{\partial u} & \dfrac{\partial y}{\partial v}
\end{pmatrix}
=
\begin{pmatrix}
\dfrac{1}{2} & \dfrac{1}{2} \\
\dfrac{1}{2} & -\dfrac{1}{2}
\end{pmatrix}.
$$

:::

::: {.fragment}
Then

$$
\det J
= \dfrac{1}{2}\cdot\left(-\dfrac{1}{2}\right)
 - \dfrac{1}{2}\cdot\dfrac{1}{2}
= -\dfrac{1}{4} - \dfrac{1}{4}
= -\dfrac{1}{2},
$$

:::

::: {.fragment}
so

$$
|\det J| = \dfrac{1}{2}.
$$

:::

---

**Step 3: Write the joint density of $(U, V)$**

::: {.fragment}
$$
f_{U,V}(u,v)
= f_{X,Y}\!\left(\dfrac{u+v}{2}, \dfrac{u-v}{2}\right)\cdot \dfrac{1}{2}.
$$

:::

::: {.fragment}

```{r}
#| fig-height: 4
#| code-fold: true
set.seed(1234)

n <- 5000
X <- rnorm(n)
Y <- rnorm(n)

U <- X + Y
V <- X - Y

par(mfrow = c(1, 2))

plot(X, Y, pch = 16, cex = 0.4, col = "steelblue",
     main = "Original (X, Y)",
     xlab = "X", ylab = "Y")

plot(U, V, pch = 16, cex = 0.4, col = "tomato",
     main = "Transformed (U = X+Y, V = X-Y)",
     xlab = "U", ylab = "V")
```

:::

- $(X,Y)$ is a round cloud (independent normals).
- $(U,V)$ becomes an axis‑aligned ellipse — the transformation **stretches and rotates** the cloud.
- This visually confirms the Jacobian‑based derivation.

---

### Example 2: Polar Coordinates

Let $(X, Y)$ have joint density $f_{X,Y}(x,y)$. Define

<span style="display: block; margin: 24px 0;"></span>

$$
U = R = \sqrt{X^2 + Y^2}, \qquad V = \Theta = \arctan(Y/X).
$$

<span style="display: block; margin: 24px 0;"></span>

The Jacobian determinant is $r$, which explains why polar integrals include the factor $r$.

We want $f_{R,\Theta}(r,\theta)$.


::: {.fragment}
**Step 1: Invert the transformation**

The inverse map is the usual polar–Cartesian relation:

$$
X = r\cos\theta, \qquad Y = r\sin\theta.
$$

:::

::: {.fragment}
So

$$
x = h_1(r,\theta) = r\cos\theta, \quad
y = h_2(r,\theta) = r\sin\theta.
$$

:::

::: {.fragment}
**Step 2: Compute the Jacobian**

$$
J =
\begin{pmatrix}
\dfrac{\partial x}{\partial r} & \dfrac{\partial x}{\partial \theta} \\
\dfrac{\partial y}{\partial r} & \dfrac{\partial y}{\partial \theta}
\end{pmatrix}
=
\begin{pmatrix}
\cos\theta & -r\sin\theta \\
\sin\theta & r\cos\theta
\end{pmatrix}.
$$

:::

::: {.fragment}
Then

$$
\det J
= \cos\theta \cdot r\cos\theta
 - (-r\sin\theta)\cdot\sin\theta
= r(\cos^2\theta + \sin^2\theta)
= r.
$$

:::

::: {.fragment}
So

$$
|\det J| = r.
$$
:::

---

**Step 3: Joint density of $(R,\Theta)$**

$$
f_{R,\Theta}(r,\theta)
= f_{X,Y}(r\cos\theta, r\sin\theta)\cdot r.
$$

::: {.fragment}

```{r}
#| code-fold: true
#| fig-height: 4
set.seed(1234)

n <- 5000
X <- rnorm(n)
Y <- rnorm(n)

R <- sqrt(X^2 + Y^2)
Theta <- atan2(Y, X)

par(mfrow = c(1, 2))

hist(R, breaks = 40, freq = FALSE, col = "lightgray",
     main = "Distribution of R",
     xlab = "R")
curve(dchisq(x^2, df = 2) * 2*x, add = TRUE, col = "darkgreen", lwd = 2)

hist(Theta, breaks = 40, freq = FALSE, col = "lightblue",
     main = "Distribution of Theta",
     xlab = expression(theta))
```

:::

- R follows the **Rayleigh distribution** (equivalently, $R^2\sim \chi_2^2$).
- $\Theta$ is uniform on $[-\pi ,\pi ]$.

---

```{r}
#| fig-height: 4
#| code-fold: true
par(mfrow = c(1, 3))

plot(X, Y, pch = 16, cex = 0.4, col = "steelblue",
     main = "Original (X, Y)",
     xlab = "X", ylab = "Y")

plot(R, Theta, pch = 16, cex = 0.4, col = "green",
     main = "Transformed (R, Theta)",
     xlab = "R", ylab = "Theta")

plot(R*cos(Theta), R*sin(Theta), pch = 16, cex = 0.4, col = "tomato",
     main = "Back to Cartesian via Polar",
     xlab = "Rcos(theta)", ylab = "Rsin(theta)")
```

- Original $(X, Y)$ shows a circular cloud centred at the origin.
- Transformed $(R, \Theta)$ shows a rectangular cloud with horizontal axis $R \ge 0$ and vertical axis $\Theta \in [-\pi, \pi].$
- Revert back to cartesian via polar using $(X,Y) = (Rcos(\Theta), Rsin(\Theta))$, then the same circular cloud reappears.

---

### Example 3: Linear Transformations

Let $(X, Y)$ have joint density $f_{X,Y}(x,y)$, and let

<span style="display: block; margin: 24px 0;"></span>

$$
\begin{pmatrix} U \\ V \end{pmatrix}
=
A \begin{pmatrix} X \\ Y \end{pmatrix},
\quad
A =
\begin{pmatrix}
a & b \\
c & d
\end{pmatrix},
$$

with $\det A \neq 0$. Then the Jacobian determinant is simply $|\det A|$.

This case is especially important for the **multivariate normal distribution**, where linear transformations **create covariance structures**. The correlated‑normal simulation you used earlier is a special case of this idea.

::: {.fragment}

**Step 1: Invert the transformation**

Because $\det A \neq 0$, $A$ is invertible:

$$
\begin{pmatrix} X \\ Y \end{pmatrix}
= A^{-1} \begin{pmatrix} U \\ V \end{pmatrix}.
$$

:::

::: {.fragment}
$$
A^{-1} =
\frac{1}{ad - bc}
\begin{pmatrix}
d & -b \\
-c & a
\end{pmatrix}.
$$

:::

::: {.fragment}
So

$$
x = h_1(u,v), \quad y = h_2(u,v)
$$

are linear functions of $(u,v)$.
:::

::: {.fragment}
**Step 2: Jacobian of the inverse**

For a linear map, the Jacobian matrix of the inverse is just $A^{-1}$, so

$$
|\det J| = |\det A^{-1}| = \frac{1}{|\det A|}.
$$

:::



---

**Step 3: Joint density of $(U,V)$**

$$
f_{U,V}(u,v)
= f_{X,Y}\big(h_1(u,v), h_2(u,v)\big)\cdot \frac{1}{|\det A|}.
$$

::: {.fragment}

```{r}
#| code-fold: true
#| fig-height: 4
set.seed(1234)

n <- 5000
X <- rnorm(n)
Y <- rnorm(n)

A <- matrix(c(1, 0.8,
              0.2, 1), nrow = 2, byrow = TRUE)

UV <- A %*% rbind(X, Y)
U <- UV[1, ]
V <- UV[2, ]

par(mfrow = c(1, 2))

plot(X, Y, pch = 16, cex = 0.4, col = "gray40",
     main = "Original (X, Y)",
     xlab = "X", ylab = "Y")

plot(U, V, pch = 16, cex = 0.4, col = "firebrick",
     main = "Transformed (U, V) = A(X, Y)",
     xlab = "U", ylab = "V")
```

:::


- The original cloud is round (independent normals).
- The transformed cloud becomes tilted and stretched, showing induced correlation.


# Part 2: Linear Trasformations and the Multivariate Normal

## Why Linear Transformation

Linear transformations are central to probability and statistics, especially for **multivariate normal distributions**.

::: {.fragment .note}
**A key idea:**

- start with **independent standard normals**
- apply a **matrix transformation**
- create **correlation and covariance structure**
:::

::: {.fragment}
This lets us construct **any multivariate normal distribution** from a standard one.
:::

::: {.fragment .tip}
**Why This Matters**

::: {.fragment}
Linear transformations underpin:
:::

- simulation of **correlated variables**
- **regression** and linear models
- **principal component analysis (PCA)**
- **Bayesian multivariate priors**
- many other multivariate methods
:::


::: {.fragment .center}
This topic is the bridge between:

**probability theory** and **practical statistical modelling**
:::

---

## Linear Transformations of Random Vectors

::: {.note .fragment}
Let

<span style="display: block; margin: 24px 0;"></span>

$$
\mathbf{X} = (X_1, X_2, \dots, X_k)^\top
$$

<span style="display: block; margin: 24px 0;"></span>

be a random vector, and let $A$ be a fixed $m \times k$ matrix. 

::: {.fragment}
A **linear transformation** of $\mathbf{X}$ is

<span style="display: block; margin: 24px 0;"></span>

$$
\boxed{\mathbf{Y} = A\mathbf{X} + \mathbf{b}}
$$

<span style="display: block; margin: 24px 0;"></span>
where $\mathbf{b}$ is a constant vector.

::: 


<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
This transformation:
:::

- rotates, stretches, or compresses the space,  
- shifts the mean by $\mathbf{b}$,  
- and reshapes the covariance structure through the matrix $A$.

:::

---

## The Multivariate Normal Distribution

::: {.tip}
**Why Multivariate Normal (MVN)?**

> It is the only multivariate distribution that is **stable under linear transformation** and **fully characterised by mean + covariance**.

That makes it mathematically tractable, computationally efficient, and an ideal building block for simulating dependence.

:::


::: {.fragment .note}
A random vector $\mathbf{X}$ is **MVN**:

$$
\mathbf{X} \sim N(\boldsymbol{\mu}, \Sigma),
$$

- $\boldsymbol{\mu}$ is the mean vector,  
- $\Sigma$ is the covariance matrix (symmetric and positive‑definite).

::: {.fragment}
$$
f_{\mathbf{X}}(\mathbf{x})
= \frac{1}{(2\pi)^{k/2} |\Sigma|^{1/2}}
\exp\!\left(
-\frac{1}{2}(\mathbf{x}-\boldsymbol{\mu})^\top
\Sigma^{-1}(\mathbf{x}-\boldsymbol{\mu})
\right).
$$

:::

::: {.fragment}
$$
\mathbf{X} \sim N(\boldsymbol{\mu}, \Sigma)
\quad \Longrightarrow \quad
A\mathbf{X} + \mathbf{b} \sim N(A\boldsymbol{\mu} + \mathbf{b},\; A\Sigma A^\top).
$$

:::

This means:

- linear transformations of normals are still normal (preserve normality),  
- the mean transforms linearly,  
- the covariance transforms quadratically.
:::

---

### Example 4: Linear Transformation of a Bivariate Normal

Simulate $\mathbf{X}\sim N(\mu ,\Sigma )$ with mean vector and covariance matrix:

$$
\mu = (1,2)^{\top}, \qquad
\Sigma =\left( \begin{matrix}4&1\\ 1&1\end{matrix}\right).
$$

The entries encode $\text{Var}(X_1) = 4$, $\text{Var}(X_2) = 1$ and $\text{Cov}(X_1,X_2) = 1$.

Then, apply $\mathbf{Y}=A\mathbf{X}+b$ with

$$
A =\left( \begin{matrix}1&0\\ -1&2\end{matrix}\right), \qquad
b = (0, 1)^{\top}
$$

```{r}
#| code-fold: true
#| fig-height: 3
#| fig-align: center
set.seed(1234)

library(MASS)

# Original distribution
mu  <- c(1, 2)
Sigma <- matrix(c(4, 1,
                  1, 1), 2, 2)

n <- 5000
X <- mvrnorm(n, mu = mu, Sigma = Sigma)

# Linear transformation
A <- matrix(c(1, 0,
             -1, 2), 2, 2)
b <- c(0, 1)

# Matrix algebra
Y <- t(A %*% t(X)) + matrix(b, n, 2, byrow = TRUE)

par(mfrow = c(1, 2), mar = c(4,4,1,1))

plot(X[,1], X[,2], pch = 16, cex = 0.4, col = "steelblue",
     main = "Original X ~ N(mu, Sigma)",
     xlab = "X1", ylab = "X2")

plot(Y[,1], Y[,2], pch = 16, cex = 0.4, col = "tomato",
     main = "Transformed Y = A X + b",
     xlab = "Y1", ylab = "Y2")
```

- The original cloud has a certain tilt and spread.
- The transformed cloud is rotated and stretched exactly as predicted by $A\Sigma A^{\top }$.

---

## Constructing a Multivariate Normal via Cholesky

- To simulate correlated variables, we need a way to impose a given covariance structure on independent normals. 
- The **Cholesky decomposition** provides a simple and efficient solution. 
- It transforms standard normal variables into correlated ones using a **matrix factorisation** of the covariance matrix. 
- This *generalises* the familiar two-dimensional construction to **higher dimensions**.

::: {.fragment .note}
If $\Sigma$ is a covariance matrix, we can write:

$$
\Sigma = LL^\top,
$$

where $L$ is the **Cholesky factor** (lower triangular). 

<span style="display: block; margin: 24px 0;"></span>


::: {.fragment}
If $\mathbf{Z} \sim N(\mathbf{0}, I)$, then

<span style="display: block; margin: 24px 0;"></span>

$$
\mathbf{X} = L\mathbf{Z} + \boldsymbol{\mu}
$$

<span style="display: block; margin: 24px 0;"></span>

has distribution $N(\boldsymbol{\mu}, \Sigma)$.
:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
This generalises earlier 2D construction:

$$
X = \rho Z_1 + \sqrt{1-\rho^2}\, Z_2,
$$

:::
<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}

which is exactly the Cholesky factor of $\begin{pmatrix}1 & \rho \\ \rho & 1\end{pmatrix}$.

:::

:::

---

### Example 5: Constructing Correlated Normals via Cholesky

Define the covariance matrix

$$\Sigma =\left( \begin{matrix}1&\rho \\ \rho &1\end{matrix}\right) =\left( \begin{matrix}1&0.8\\ 0.8&1\end{matrix}\right).$$

::: {.fragment}

```{r}
#| code-fold: true
#| fig-height: 3.5
set.seed(1234)

n <- 5000
rho <- 0.8

# Covariance matrix
Sigma <- matrix(c(1, rho,
                  rho, 1), 2, 2)

# Cholesky factor: L^T L (upper triangular)
L <- chol(Sigma)

# Generate 2n independent N(0,1) values
Z <- matrix(rnorm(2*n), n, 2)

# Construct X = Z L^T (linear transformation)
X <- Z %*% L

par(mfrow = c(1, 2), mar = c(4,4,1,1))

plot(Z[,1], Z[,2], pch = 16, cex = 0.4, col = "gray60",
     main = "Z ~ N(0, I)",
     xlab = "Z1", ylab = "Z2")

plot(X[,1], X[,2], pch = 16, cex = 0.4, col = "firebrick",
     main = "X = Z L^T (Correlated Normals)",
     xlab = "X1", ylab = "X2")
```

:::

- $Z$ is a round cloud (independent normals).
- $X$ becomes a tilted ellipse with correlation $\rho =0.8$.
- This is how Cholesky induces covariance.

---

### Example 6: General k‑Dimensional Construction

Simulate $\mathbf{X}\sim N(\mu ,\Sigma )$ using Cholesky. Define a mean vector and a covariance matrix:

$$
\mu =(0,\; 1,\; 2)^{\top }, \qquad
\Sigma =\left( \begin{matrix}1&0.5&0.2\\ 0.5&2&0.3\\ 0.2&0.3&1\end{matrix}\right).
$$

::: {.fragment}
```{r}
set.seed(1234)
mu <- c(0, 1, 2)
Sigma <- matrix(c(1, 0.5, 0.2,
                  0.5, 2, 0.3,
                  0.2, 0.3, 1), 3, 3)

n <- 5000
L <- chol(Sigma) # Cholesky factor L^T
Z <- matrix(rnorm(3*n), n, 3) # Standard normals

# Construct X = Z L^T + mu
X <- Z %*% L + matrix(mu, n, 3, byrow = TRUE)

round(cov(X), 2) # sample covariance
```

:::

- The sample covariance of $X$ matches $\Sigma$, suggesting the construction works in any dimension.

---

```{r}
#| code-fold: true
#| fig-height: 5
# 3D scatterplot
library(scatterplot3d)

scatterplot3d(X[,1], X[,2], X[,3],
              pch = 16, cex.symbols = 0.4,
              color = "steelblue",
              main = "3D Scatterplot of Multivariate Normal Sample",
              xlab = "X1", ylab = "X2", zlab = "X3")
```

- A 3D elliptical cloud whose shape reflects the covariance matrix.
- The cloud is tilted and stretched in directions determined by $\Sigma$.
- This visually reinforces the idea that $X=LZ+\mu$.


# Part 3: Copulas

## Why Copulas

So far, we have constructed dependence using linear transformations and Cholesky decomposition. 

This approach is powerful, but it has an important limitation:

::: {.fragment}
> It ties dependence to **normality**.
:::


::: {.fragment}
In many real-world applications (finance, insurance, environmental modelling), dependence can be:
:::

- non-linear,
- asymmetric,
- or stronger in extremes (tail dependence).


::: {.fragment}
To model such behaviour, we need a more flexible framework.

This leads to the idea of **copulas**.
:::

<br>

::: {.fragment .block .green}
A **copula** is a function that allows us to construct a joint distribution by combining:

- marginal distributions, and
- a dependence structure.
:::

---

## Copula

::: {.note}
**Key Idea**

::: {.fragment}
Any joint distribution can be decomposed as:

$$
F_{X,Y}(x,y) = C\big(F_X(x), F_Y(y)\big),
$$

where:
:::

- $F_X, F_Y$ are marginal CDFs,
- $C(u,v)$ is a copula.

::: {.fragment}
This result is known as **Sklar’s Theorem**.
:::

::: {.fragment}
**Interpretation**
:::

- Marginals describe individual behaviour.
- The copula describes dependence.

:::

::: {.tip .fragment}
**Why Copulas Matter in Simulation**

::: {.fragment}
Copulas allow us to:
:::

- simulate dependence without assuming normality,
- combine any marginal distributions (e.g., Normal + Exponential),
- model tail dependence (important in risk modelling),
- separate modelling into marginals (easy) or dependence (flexible).


:::

---

## Gaussian Copula {.unnumbered}

The most commonly used copula in simulation is the Gaussian copula.

<br>

#### Simulation Algorithm


::: {.fragment .block .green}
1.	Choose:
	- marginal distributions $F_X, F_Y$,
	- correlation parameter $\rho$

:::

::: {.fragment .block .yellow}
2.	Construct covariance matrix $\hspace{7em} \Sigma =\begin{pmatrix}1 & \rho \\\rho & 1 \end{pmatrix}$

:::

::: {.fragment .block .blue}
3.	Generate $\hspace{13em} (Z_1, Z_2) \sim N(0, \Sigma)$

:::

::: {.fragment .block .purple}

4.	Transform $\hspace{14em} U_i = \Phi(Z_i)$

:::

::: {.fragment .block .grey}
5.	Apply inverse CDF $\hspace{7em} X = F_X^{-1}(U_1), \quad Y = F_Y^{-1}(U_2)$

:::


---

### Example 7: Normal–Exponential Dependence

Simulate $X \sim N(0,1)$ and $Y \sim \text{Exp}(1)$ with dependence $\rho = 0.7$

```{r}
#| code-fold: true
#| fig-height: 4
set.seed(1234)

library(MASS)

n <- 5000
rho <- 0.7

# Step 1: correlated normals
Sigma <- matrix(c(1, rho,
                  rho, 1), 2, 2)

Z <- mvrnorm(n, mu = c(0,0), Sigma = Sigma)

# Step 2: convert to uniforms
U <- pnorm(Z)

# Step 3: apply inverse CDFs
X <- qnorm(U[,1])      # Normal
Y <- qexp(U[,2])       # Exponential

plot(X, Y, pch = 16, cex = 0.4, col = "steelblue",
     main = "Gaussian Copula (Normal + Exponential)",
     xlab = "X ~ N(0,1)", ylab = "Y ~ Exp(1)")
```


- Marginals are preserved exactly,
- Dependence is induced via the copula,
- The shape is not elliptical (unlike MVN).

---

### Limitations

The Gaussian is **weak in modelling extreme co-movements.** Alternatively, we can use t-copula to capture tail dependence, Clayton copula to capture lower tail dependence, or Gumbel copula to capture upper tail dependence.

### Simulation of Dependent Variables (Summary) {.fragment}


| | Linear Transformation | Copula-Based Simulation |
|----------|-----------------------|-----------------------|
| Use | $X=LZ+\mu$ | $U_i = \Phi(Z_i), \quad X_i = F_i^{-1}(U_i)$ |
| Works for | Normal only | Any marginals |
| Dependence | Linear | Flexible |
| Pros | Simple and efficient | Very flexible, realistic modelling |
| Cons | Limited flexibility | Slightly more complex |
| Use case | Statistics, regression | Finance, risk, insurance |

: {.striped .hover .fragment}

<br>

::: {.note .fragment}
**General Algorithm**

1.	Specify marginals $F_1, \dots, F_k$
2.	Choose dependence structure: covariance matrix (MVN) or copula
3.	Generate base variables: $Z \sim N(0,I)$
4.	Induce dependence via $LZ$ (Cholesky) or copula
5.	Transform to desired marginals
:::



# 📈 Case Study: Portfolio Risk Simulation

---

#### 📈 Case Study: Portfolio Risk Simulation

In this example, we study the **distribution of returns**, which describes how much an asset or portfolio gains or loses over time (e.g., daily returns). Understanding this distribution helps answer questions such as:

- How large can losses be?
- How likely are extreme events?
- How does dependence between assets affect risk?

::: {.fragment}
**Financial theory** assumes that returns follow **a normal (Gaussian) or lognormal distribution**, which means:
:::

- most outcomes are close to the average (symmetric around the mean),
- extreme gains or losses are rare,
- dependence between assets is captured using correlation.

::: {.fragment}
```{r}
#| echo: false
#| fig-height: 2.8
#| fig-align: center
set.seed(123)  # For reproducibility
data <- rnorm(500, mean = 0, sd = .01)

par(mar = c(4,4,1,1))

# Plot histogram with density curve
hist(
  data,
  breaks = 30,                # Number of bins
  probability = TRUE,         # Show density instead of counts
  col = "lightblue",
  border = "white",
  main = "Distribution of Returns",
  xlab = "Percentage Returns"
)

# Add density curve
lines(density(data), col = "red", lwd = 2)

# Add theoretical normal curve for comparison
x <- seq(min(data), max(data), length = 100)
y <- dnorm(x, mean = 0, sd = .01)
lines(x, y, col = "darkgreen", lwd = 2, lty = 2)

# Add legend
legend(
  "topright",
  legend = c("Real Density", "Theoretical Normal"),
  col = c("red", "darkgreen"),
  lwd = 2,
  lty = c(1, 2)
)
```

:::

---

In practice, returns often behave differently:

- **Fat tails:** Extreme events occur more often than predicted by a normal distribution.
- **High peaks (leptokurtosis):** Returns may be more concentrated around the mean.
- **Skewness:** Distributions may be asymmetric, with markets reacting differently to positive versus negative events (overreaction or underreaction).
- **Dependency:** Dependence between assets can become stronger during extreme events.

::: {.fragment}
This means simple normal models may underestimate risk, especially for large losses.
:::

```{r}
#| echo: false
#| fig-height: 4
#| fig-align: center
set.seed(123)  # For reproducibility
data <- rt(400, df=5)

# Plot histogram with density curve
hist(
  data,
  breaks = 30,                # Number of bins
  probability = TRUE,         # Show density instead of counts
  col = "lightblue",
  border = "white",
  main = "Distribution of Returns",
  xlab = "Percentage Returns",
  xaxt = "n"   # suppress default x-axis
)

# Custom percentage x-axis
ticks <- pretty(range(data))
axis(
  side = 1,
  at = ticks,
  labels = paste0(ticks / 100) # paste0(ticks * 100, "%")
)

# Add density curve
lines(density(data), col = "red", lwd = 2)

# Add theoretical t curve for comparison
x <- seq(min(data), max(data), length = 100)
y <- dt(x, df=5)
lines(x, y, col = "darkgreen", lwd = 2, lty = 2)

# Add legend
legend(
  "topleft",
  legend = c("Real Density", "Theoretical t-distribution"),
  col = c("red", "darkgreen"),
  lwd = 2,
  lty = c(1, 2)
)
```


---

## Problem Statement

::: {.example}
A financial analyst wants to **model the joint daily returns** of two assets in a portfolio, **Asset A** and **Asset B**, with equal weights of 50%. 

The analyst wants to estimate:

- the mean and standard deviation of the portfolio return
- the 1% and 5% Value-at-Risk (VAR), and 
- the probability of a large one-day portfolio loss.

::: {.fragment}
Since the asset returns are dependent, simulating each asset separately would underestimate the risk of simultaneous losses. 

Therefore, the analyst considers three modelling approaches:
:::

- a **multivariate normal model** as a baseline for correlated returns,
- a **copula-based model** to allow more flexible marginal distributions and dependence, especially under extreme market conditions, and
- an **independent model** as a benchmark.

::: {.fragment}
The analyst then conduct data analysis to understand the distribution of returns for Asset A and Asset B.
:::

:::

---

### Multivariate Normal Model

The portfolio weights are

$$
w_1 = 0.5, \qquad w_2 = 0.5,
$$

so the portfolio return is

$$
R_p = 0.5R_1 + 0.5R_2.
$$

Based on the data analysis, the analyst assumes:

- $R_1$ has a Normal marginal distribution, $R_1 \sim N(0.0005, 0.015^2)$.
- $R_2$ has a Normal marginal distribution, $R_2 \sim N(0.0008, 0.02^2)$.
- Dependence is introduced via a covariance matrix with correlation $\rho = 0.6$.

::: {.fragment}
```{r}
library(MASS)
set.seed(1234)
n <- 10000
rho <- 0.6 # correlation
w <- c(0.5, 0.5) # Portfolio weights
mu <- c(0.0005, 0.0008) # Mean vector (daily returns)
sd1 <- 0.015 # standard deviation of R1
sd2 <- 0.02  # standard deviation of R2
Sigma <- matrix(c(sd1^2, rho*sd1*sd2,
                  rho*sd1*sd2, sd2^2), 2, 2) # Covariance matrix
R_gauss <- mvrnorm(n, mu = mu, Sigma = Sigma) # Simulate correlated normal returns
port_gauss <- R_gauss %*% w # Portfolio return
```
:::

---

```{r}
#| code-fold: true
risk_summary <- c(
  Mean = mean(port_gauss), # mean of portfolio return
  SD = sd(port_gauss),     # standard devation of portfolio return
  VaR_5 = quantile(port_gauss, 0.05), # 5% Value-at-Risk
  VaR_1 = quantile(port_gauss, 0.01), # 1% Value-at-Risk
  Prob_Loss_gt_3pct = mean(port_gauss < -0.03) # probability of losing more than 3% in one day
)

library(knitr)
kable(round(risk_summary, 4))
```

```{r}
#| code-fold: true
#| fig-align: center
#| fig-height: 3
par(mfrow = c(1, 2), mar = c(4,4,1,1))

plot(R_gauss[,1], R_gauss[,2],
     pch = 16, cex = 0.4, col = "steelblue",
     main = "Gaussian Model",
     xlab = "Asset A return", ylab = "Asset B return")

hist(port_gauss, breaks = 40, freq = FALSE, col = "lightgray",
     main = "Portfolio Return (Gaussian)",
     xlab = "Portfolio return")
abline(v = -0.03, col = "red", lwd = 2, lty = 2)
```

---

### Copula-based Model

The portfolio weights and portfolio return are

$$
w_1 = 0.5, \qquad w_2 = 0.5, \qquad R_p = 0.5R_1 + 0.5R_2.
$$


Based on the data analysis, the analyst assumes:

- $R_1$ has a Normal marginal distribution, $R_1 \sim N(0.0005, 0.015^2)$.
- $R_2$ has a heavier-tailed $t$ marginal distribution with 4 degrees of freedom, rescaled to have approximately standard deviation $0.02$.
- Dependence is introduced using a Gaussian copula with correlation $\rho = 0.6$.

::: {.fragment}
```{r}
set.seed(1234)
n <- 10000
rho <- 0.6  # Correlation
Sigma_cop <- matrix(c(1, rho, rho, 1), 2, 2)
Z <- mvrnorm(n, mu = c(0, 0), Sigma = Sigma_cop) 
U <- pnorm(Z) # Convert to Uniform

# Asset A: Normal marginal
X1 <- qnorm(U[,1], mean = 0.0005, sd = 0.015)

# Asset B: heavy-tailed t marginal
# Standard t(df=4) has variance df/(df-2) = 2, so divide by sqrt(2)
# to approximately standardise before scaling
X2 <- 0.0008 + 0.02 * (qt(U[,2], df = 4) / sqrt(2))

R_copula <- cbind(X1, X2)
port_copula <- R_copula %*% w # Portfolio return
```

:::

---

```{r}
#| code-fold: true
risk_summary2 <- c(
  Mean = mean(port_copula), # mean of portfolio return
  SD = sd(port_copula),     # standard devation of portfolio return
  VaR_5 = quantile(port_copula, 0.05), # 5% Value-at-Risk
  VaR_1 = quantile(port_copula, 0.01), # 1% Value-at-Risk
  Prob_Loss_gt_3pct = mean(port_copula < -0.03) # probability of losing more than 3% in one day
)

library(knitr)
kable(round(risk_summary2, 4))
```

```{r}
#| code-fold: true
#| fig-align: center
#| fig-height: 3
par(mfrow = c(1, 2))

plot(R_copula[,1], R_copula[,2],
     pch = 16, cex = 0.4, col = "tomato",
     main = "Gaussian Copula Model",
     xlab = "Asset A return", ylab = "Asset B return")

hist(port_copula, breaks = 40, freq = FALSE, col = "lightblue",
     main = "Portfolio Return (Copula)",
     xlab = "Portfolio return")
abline(v = -0.03, col = "red", lwd = 2, lty = 2)
```

---

### Independent Model (Benchmark)

Using the same model setup as Copula-based model, we simulate returns of Asset A and Asset B independently:

```{r}
set.seed(1234)

R1_ind <- rnorm(n, mean = 0.0005, sd = 0.015)
R2_ind <- 0.0008 + 0.02 * (rt(n, df = 4) / sqrt(2))

R_port_ind <- 0.5 * R1_ind + 0.5 * R2_ind
```

```{r}
#| code-fold: true
#| code-summary: Comparison of Scatterplots
#| fig-height: 2.5
#| fig-align: center
par(mfrow = c(1, 3), mar = c(4,4,1,1))

plot(R1_ind, R2_ind,
     pch = 16, cex = 0.4, col = "palegreen",
     main = "Independent Model",
     xlab = "Asset A return", ylab = "Asset B return")

plot(R_gauss[,1], R_gauss[,2],
     pch = 16, cex = 0.4, col = "steelblue",
     main = "Gaussian Model",
     xlab = "Asset A return", ylab = "Asset B return")

plot(R_copula[,1], R_copula[,2],
     pch = 16, cex = 0.4, col = "tomato",
     main = "Gaussian Copula Model",
     xlab = "Asset A return", ylab = "Asset B return")

```

- **Independent model**
→ circular cloud → no structure
→ returns move independently
- **Gaussian model**
→ elliptical cloud
→ linear dependence
- **Copula model**
→ similar dependence, but more extreme points
→ heavier tails visible

---

```{r}
#| code-fold: true
#| code-summary: Comparison of Marginal Histograms
#| fig-height: 2.5
#| fig-align: center
par(mfrow = c(1, 3), mar = c(4,4,1,1))

hist(R_port_ind, breaks = 40, freq = FALSE, col = "lightpink",
     main = "Portfolio Return (Independent)",
     xlab = "Portfolio return")
abline(v = -0.03, col = "red", lwd = 2, lty = 2)

hist(port_gauss, breaks = 40, freq = FALSE, col = "lightgray",
     main = "Portfolio Return (Gaussian)",
     xlab = "Portfolio return")
abline(v = -0.03, col = "red", lwd = 2, lty = 2)

hist(port_copula, breaks = 40, freq = FALSE, col = "lightblue",
     main = "Portfolio Return (Copula)",
     xlab = "Portfolio return")
abline(v = -0.03, col = "red", lwd = 2, lty = 2)
```


- **Independent** → narrow distribution &emsp;&emsp;&emsp; **Gaussian** → wider spread &emsp;&emsp;&emsp; **Copula** → similar spread, fatter tails

<br>

::: {.fragment}
```{r}
#| echo: false
comparison <- rbind(
  Gaussian = c(
    Mean = mean(port_gauss),
    SD = sd(port_gauss),
    VaR_5 = quantile(port_gauss, 0.05),
    VaR_1 = quantile(port_gauss, 0.01),
    Prob_Loss_gt_3pct = mean(port_gauss < -0.03)
  ),
  Copula = c(
    Mean = mean(port_copula),
    SD = sd(port_copula),
    VaR_5 = quantile(port_copula, 0.05),
    VaR_1 = quantile(port_copula, 0.01),
    Prob_Loss_gt_3pct = mean(port_copula < -0.03)
  ),
  Independent = c(
    Mean = mean(R_port_ind),
    SD = sd(R_port_ind),
    VaR_5 = quantile(R_port_ind, 0.05),
    VaR_1 = quantile(R_port_ind, 0.01),
    Prob_Loss_gt_3pct = mean(R_port_ind < -0.03)
  )
)

kable(round(comparison, 4))
```

:::

<br>

- The means are very similar → expected return is not sensitive to dependence.
- The standard deviation increases significantly when dependence is introduced.
- Dependence roughly doubles the probability of large losses.


---

::: {.note}
**Value-at-Risk (VaR)** is a commonly used measure of downside risk (losses or left tail of the distribution). It measures **“How bad can losses be, with a given level of confidence?”**

::: {.fragment}
Formally, the VaR at level $\alpha$ (e.g., 5% or 1%) is the threshold such that:

$$
P(R_p \leq \text{VaR}_\alpha) = \alpha,
$$

where $R_p$ is the portfolio return.
:::

- VaR (5%) = −0.02 means there is a 5% chance that the portfolio return is worse than −2%.
- VaR (1%) = −0.04 means there is a 1% chance that the loss exceeds 4%.
:::

<br>

::: {.fragment}

```{r}
#| echo: false
comparison <- rbind(
  Gaussian = c(
    Mean = mean(port_gauss),
    SD = sd(port_gauss),
    VaR_5 = quantile(port_gauss, 0.05),
    VaR_1 = quantile(port_gauss, 0.01),
    Prob_Loss_gt_3pct = mean(port_gauss < -0.03)
  ),
  Copula = c(
    Mean = mean(port_copula),
    SD = sd(port_copula),
    VaR_5 = quantile(port_copula, 0.05),
    VaR_1 = quantile(port_copula, 0.01),
    Prob_Loss_gt_3pct = mean(port_copula < -0.03)
  ),
  Independent = c(
    Mean = mean(R_port_ind),
    SD = sd(R_port_ind),
    VaR_5 = quantile(R_port_ind, 0.05),
    VaR_1 = quantile(R_port_ind, 0.01),
    Prob_Loss_gt_3pct = mean(R_port_ind < -0.03)
  )
)

kable(round(comparison, 4))
```

:::

<br>

- Both dependent models produce more negative VaR than the independent case.
- The copula model has the worst 1% VaR.

::: {.fragment}
> **Takeaway:** Ignoring dependence underestimates risk.
> Heavy tails (copula model) increase extreme losses.
:::

