---
title: "STAT2005 Computer Simulation"
subtitle: "Lecture 4 — Simulation of Discrete Random Variables"
author: 
  - name: Sam Wiwatanapataphee (275287a@curtin.edu.au)
date: "2026-03-12"
date-format: "D MMM YYYY"
format:
  revealjs:
    html-math-method: mathjax
    theme: [simple, custom.scss]
    logo: figs/curtin-logo2.png
    margin: 0.05
    center: false
    slide-number: true
    incremental: true
    toc: false
    code-line-numbers: true
    code-copy: true
    scrollable: true
    title-slide: false
    embed-resources: false # true to export as single html: not compatible with chalkboard
    # scroll-view: true
    menu:
      titleSelector: 'h1, h2, h3'
      useTextContentForMissingTitles: false 
      hideMissingTitles: true
    chalkboard: true
    # Use this if you want a wider slide layout:
    # width: 1280 (default is 1050)
    # height: 720 (default is 700)
execute:
  echo: true
  warning: false
  message: false
# jupyter: python3 # cannot have both r and python
---


## Goals today

:::: {.columns}
::: {.column}
- **Part 1:** Statistical Inference (continue)
    - Review: Binomial and Beta Distributions
    - Classical Inference for Proportions
    - Alternative Classical Methods for Proportion Inference
    - Bayesian Inference about a Population Proportion
    - Types of Priors
      - Non-informative
      - Expert Priors (Informative Priors)
      - Conjugate Priors
:::
:::{.column}
- **Part 2:** Simulation of Discrete Random Variables
    - The precision method
    - Discrete inverse CDF method
    - Case study: Epidemic SIR Modelling

:::
::::

# Part 1: Statistical Inference (continue)

##  Binomial and Beta Distributions

::: {.note}

When we observe $y$ successes and $(n-y)$ failures in an experiment, we use a <span style="color: blue;">**discrete Binomial pmf**</span> to estimate the probability of success $\theta, 0 \leq \theta \leq 1,$ that is

<span style="display: block; margin: 24px 0;"></span>

$$
Y \sim \text{Binomial}(n, \theta)
$$

such that
<span style="display: block; margin: 24px 0;"></span>

$$
p(y \mid \theta) = C_y^n\theta^y (1-\theta)^{n-y}, \quad y = 0, 1, \dots, n.
$$

We know that
<span style="display: block; margin: 24px 0;"></span>

$$
E(Y \mid \theta) = n\theta, \quad \text{Var}(Y\mid\theta) = n\theta(1-\theta).
$$

:::

::: {.tip .fragment}
The <span style="color: green;">**beta distribution**</span> is a family of continuous distributions on the interval $[0,1],$ It has two positive shape parameters, $\alpha$ and $\beta$, which act as exponents and determine the distribution’s shape. Its pdf is

<span style="display: block; margin: 24px 0;"></span>

$$
\theta \sim \text{Beta}(\alpha,\beta)
$$

such that


$$
p(\theta) = \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\theta^{\alpha-1}(1-\theta)^{\beta-1}, \quad \alpha,\beta > 0, \quad 0\leq\theta\leq 1.
$$

The mean and variance are

$$
E(\theta) = \frac{\alpha}{\alpha+\beta}, \quad \text{Var}(\theta)=\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}
$$

:::

---

#### Beta Distributions

```{r}
#| fig-align: center
theta <- seq(0,1, by=0.001)
par(mfrow=c(2,2), mar=c(3,4,1,1))
plot(dbeta(theta, 1, 1),type="l", main="alpha=beta=1, mean=0.5, var=1/12")
plot(dbeta(theta, 5, 1),type="l", main="alpha=5,beta=1, mean=1/3, var=4/45")
plot(dbeta(theta, 1, 3),type="l",main="alpha=2, beta=3")
plot(dbeta(theta, 2, 5),type="l",main="alpha=1, beta=5")
```

---

## Classical Inference for Proportions

Classical inference for a binomial proportion is often presented as simple, but the usual methods perform poorly, especially when $y=0$ or $y=n.$ For a binomial model

<span style="display: block; margin: 24px 0;"></span>

$$
Y \sim \text{Binomial}(n,p),
$$

the classical estimator is

$$
\hat{p} = \frac{y}{n}.
$$

::: {.fragment}
A common large-sample $95\%$ confidence interval uses the normal approximation:
<span style="display: block; margin: 24px 0;"></span>

$$
\hat{p} \pm 1.96 \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}.
$$


This relies on the **central limit theorem**, which requires large $n$ so that

$$
Y \approx \text{Normal}\big(np,\; np(1-p)\big),
\qquad
\hat{p} \approx \text{Normal}\!\left(p,\; \frac{p(1-p)}{n}\right).
$$

In practice, the approximation may be poor for finite samples. 

:::

::: {.tip .fragment}

A common **rule of thumb** is to use the interval only when

$$
np > 5 \quad \text{and} \quad n(1-p) > 5,
$$

but this depends on the unknown $p$ rather than the estimate $\hat{p}$.

Thus, even from a classical perspective, the standard normal-approximation interval has several limitations.

:::


---

## Alternative Classical Methods for Proportion Inference

<span style="display: block; margin: 24px 0;"></span>

Classical inference typically substitutes $\hat{p}$ for the unknown proportion $p$
without adjusting for the uncertainty introduced by estimation. This can lead to
poor performance, especially in small samples.

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}

Over the years, more than twenty alternative classical intervals have been
proposed, and there is still no consensus on a single preferred method. One
widely used improvement is the **Agresti--Coull interval**.

:::

<span style="display: block; margin: 24px 0;"></span>


### Agresti--Coull Adjustment {.fragment}

<span style="display: block; margin: 24px 0;"></span>

::: {.note .fragment}

Define the adjusted estimator

<span style="display: block; margin: 24px 0;"></span>

$$
\tilde{p} = \frac{y + 2}{n + 4}.
$$

A $95\%$ confidence interval is then

<span style="display: block; margin: 24px 0;"></span>

$$
\tilde{p} \pm 1.96 \sqrt{\frac{\tilde{p}(1 - \tilde{p})}{\,n + 4\,}}.
$$


<span style="display: block; margin: 24px 0;"></span>

This method effectively adds two successes and two failures to the data, which
is analogous to using a $\text{Beta}(3,3)$ prior in a Bayesian framework.

:::

---

## Bayesian Inference about a Population Proportion

- A common inference problem is estimating a population proportion based on the number of successes and failures observed in a **binomial sample**.
- The posterior distribution represents our updated uncertainty about the proportion after combining:
  - the observed data (likelihood), and
  - prior beliefs about the parameter.
-  More formally: $\qquad \qquad \qquad \qquad \qquad \; P(\theta \mid \text{data}) \propto P(\text{data} \mid \theta)\,P(\theta).$



#### Choosing a Prior {.fragment}

::: {.fragment}
When specifying a prior, we ask:
:::

- Do we have **prior knowledge** about the parameter?
- How **strong** should that prior information be?
- What **distribution** appropriately represents those beliefs?

#### Types of Priors {.center .fragment}

:::: {.columns}
::: {.column .center .block .blue .fragment width=34%}
**Conjugate**<br>
chosen so the posterior has the same distributional family as the prior
:::
::: {.column .center .block .yellow .fragment width=28%}
**Non-informative**<br>
represent minimal prior knowledge
:::
::: {.column .center .block .red .fragment width=28%}
**Informative**<br>
incorporate *expert* knowledge or previous studies
:::
::::

- Other common prior categories include weakly informative priors, hierarchical priors, and improper priors.


---

::: {.example .fragment}

🏀 **Example: Basketball** 

Suppose we want to learn about the true free‑throw success rate $p$ of a professional basketball player.

- When we watch a game and record the number of made and missed free throws, we collect information about the player’s underlying success rate $p.$
- Assuming each free throw is independent and has the same probability of success $p,$ the number of made shots $Y$ can be modelled with a **binomial distribution**:

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
$$
Y \mid p \sim \text{Binomial}(n, p).
$$
:::

<span style="display: block; margin: 24px 0;"></span>

- Suppose we observe the player taking **60 free throws**, of which **39 are made and 21 are missed**. So:

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
$$
n = 60, \quad y = 39, \quad (n-y) = 21.
$$

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}
The **binomial likelihood** is

<span style="display: block; margin: 24px 0;"></span>

$$
P(Y= 39 \mid p) = C_{39}^{60} p^{39} (1-p)^{21}.
$$

:::

<span style="display: block; margin: 24px 0;"></span>

- This observed data will allow us to update our beliefs about the player’s true free‑throw percentage.
:::

::: {.center .fragment style="margin-top: 20px; font-size: 1.2em; color: blue;"}
**What prior should we choose for $p?$**
:::


---

## Non-informative (Flat) Priors

- Non-informative priors aim to have **minimal influence on the posterior relative to the likelihood.**
- They attempt to represent **prior ignorance or neutrality** about the parameter.
- Uniform or flat prior, where all parameter values equally likely: $p \sim \text{Uniform(0,1)}.$

::: {.example .fragment}

🏀 **Example: Basketball with Uniform or flat prior** 

- Suppose we know nothing about the player or what his shooting percentage might be.
- A simple choice is a uniform prior, where every value of $p$ between 0 and 1 is considered equally likely.
:::

::: {.fragment}
```{r}
#| code-fold: true
#| fig-height: 3.4
p <- seq(0, 1, length = 100)

prior <- rep(1, length(p))
likelihood <- dbinom(39, 60, p)
norm <- sum(prior * likelihood) # normalising constant
posterior <- prior * likelihood / norm

par(mfrow = c(3,1),
    mar = c(0,4,1,1),   # remove bottom margin
    oma = c(4,0,0,0))   # outer margin for shared x-label

plot(p, prior, type="l", xaxt="n", ylab="Prior", col="green")
plot(p, likelihood, type="l", xaxt="n", ylab="Likelihood", col="blue")
plot(p, posterior, type="l", ylab="Posterior", xlab="p", col="red")

mtext("Free throw success probability (p)", side=1, outer=TRUE, line=3)
```

:::


---

## Informative Priors

- In many situations we have prior knowledge or expert beliefs about the parameter $p.$

::: {.example .fragment}

🏀 **Example: Basketball with Informative Prior** 

- Professional players typically make between **50% and 95%** of their free throws.  
- Based on a basketball expert's knowledge, we have the following **discrete prior**.
:::

::: {.fragment}
| Free Throw Probability (p) |0.50|0.55|0.60|0.65|0.70|**0.75**|0.80|0.85|0.90|0.95|
|----------|---|---|---|---|---|---|---|---|---|---|
| Prior Probability |0.005|0.01|0.05|0.10|0.22|<span style="color: crimson;">**0.33**</span>|0.22|0.05|0.014|0.001|
:::


::: {.fragment}
```{r}
#| code-fold: true
#| fig-height: 3.3
p <- seq(0.5, 0.95, length = 10)
prior <- c(0.005, 0.01, 0.05, 0.1, 0.22, 0.33, 0.22, 0.05, 0.014, 0.001) # given by an expert

likelihood <- dbinom(39, 60, p)
norm <- sum(prior * likelihood) # normalising constant from the law of total probability
posterior <- prior * likelihood / norm

# Prior summaries
prior_mode <- p[which.max(prior)]
prior_mean <- sum(p * prior)

# Posterior summaries
posterior_mode <- p[which.max(posterior)]   # MAP estimate on this grid
posterior_mean <- sum(p * posterior)

# 98% equal-tail credible interval on the discrete grid
post_cdf <- cumsum(posterior)
lower_98 <- p[min(which(post_cdf >= 0.01))]
upper_98 <- p[min(which(post_cdf >= 0.99))]

# Plot
par(mfrow = c(3,1),
    mar = c(0,4,1,1),   # remove bottom margin
    oma = c(4,0,0,0))   # outer margin for shared x-label

plot(p, prior, type="l", xaxt="n", ylab="Prior", col="green")
abline(v = prior_mode, lty = 2, col = "darkgreen")
plot(p, likelihood, type="l", xaxt="n", ylab="Likelihood", col="blue")
plot(p, posterior, type="l", ylab="Posterior", xlab="p", col="red")
abline(v = posterior_mode, lty = 2, col = "darkred")
abline(v = c(lower_98, upper_98), lty = 3, col = "gray40")

mtext("Free throw success probability (p)", side=1, outer=TRUE, line=3)
```

:::

---

#### Informative Priors (continue)

```{r}
# Prior summaries
prior_mode <- p[which.max(prior)] # p where the largest prior probability occurs at
prior_mean <- sum(p * prior)

# Posterior summaries
posterior_mode <- p[which.max(posterior)]
posterior_mean <- sum(p * posterior)

# 98% equal-tail credible interval on the discrete grid
post_cdf <- cumsum(posterior) # cumulative posterior
lower_98 <- p[min(which(post_cdf >= 0.01))] # find the first value where CDF exceeds 0.01
upper_98 <- p[min(which(post_cdf >= 0.99))] # find the first value where CDF exceeds 0.99
```

::: {.fragment}
```{r}
cat("Prior mode =", prior_mode, ", and Prior mean =", round(prior_mean, 3))
```

:::

::: {.fragment}
```{r}
cat("Posterior mode =", posterior_mode, ", and Posterior mean =", round(posterior_mean, 3))
```

:::

- The posterior mode (also called the MAP estimate) is the value of $p$ with the largest posterior probability.
- The new expert prior has inflated the maximum posterior estimate to be 0.7.

::: {.fragment}
```{r}
cat("98% credible interval = (", lower_98, ", ", upper_98, ")\n", sep = "")
```

:::

- Given the data and the prior, there is 98% posterior probability that the true free-throw success rate $p$ lies within this interval.

---

## Beta-Binomial Conjugate Prior

- In practice, the possible values for a population proportion $p$ will usually be on a continuous scale in the region $(0,1).$ 
- Naturally, Bayes’ rule also applies to continuous distributions. 
- In this case, we can use **Beta distribution for a prior distribution** for $p.$

::: {.fragment}
$$
p \sim \text{Beta}(\alpha,\beta),
$$

with density

$$
f(p)=\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}p^{\alpha-1}(1-p)^{\beta-1}, \quad \alpha,\beta > 0, \quad 0\leq p\leq 1.
$$

<span style="display: block; margin: 24px 0;"></span>

:::

- The Beta distribution is flexible and can be interpreted as follows:
  - $\alpha - 1$ acts like prior successes,
  - $\beta - 1$ acts like prior failures,
  - the prior mean is $E[p] = \dfrac{\alpha}{\alpha+\beta},$ and
  - its mode is $\text{Mode}(t)=\dfrac{\alpha-1}{\alpha+\beta-2}, \alpha, \beta > 1.$

- Now, what would be an appropriate value for parameter $\alpha$ and $\beta?$


---

::: {.example .fragment}

🏀 **Example: Basketball with a Beta-Binomial Conjugate Prior** 

- Suppose expert knowledge suggests this player is probably around a **75% free-throw shooter**, and that this belief is worth about **20 prior shots**.

- A convenient prior is $\qquad \qquad \qquad \qquad p \sim \text{Beta}(15,5).$

- This prior has $\qquad \qquad E[p] = \dfrac{15}{20} = 0.75, \qquad \text{Mode}(p) = \dfrac{14}{18} \approx 0.778.$


::: {.fragment}
Using Bayes’ theorem and substituting the Binomial likelihood and Beta prior,

<span style="display: block; margin: 24px 0;"></span>

$$
\begin{aligned}
f(p\mid x) &\propto P(X=x\mid p)\,f(p) \\
&\propto p^{39}(1-p)^{21}\cdot p^{15-1}(1-p)^{5-1} \\
&\propto p^{53}(1-p)^{25}
\end{aligned}
$$

So the posterior is $\qquad \qquad \;\; p\mid x \sim \text{Beta}(15+39,5+21)=\text{Beta}(54,26).$
<span style="display: block; margin: 24px 0;"></span>

:::

::: {.fragment}
The posterior mean and mode are

<span style="display: block; margin: 24px 0;"></span>

$$
E[p\mid x]=\frac{54}{54+26}=\frac{54}{80}=0.675, \qquad \text{Mode}(p)=\frac{54-1}{54+26-2}=\frac{53}{78}\approx 0.679.
$$

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}

For comparison, the sample proportion is $\quad\hat p=\dfrac{39}{60}=0.65.$

<span style="display: block; margin: 24px 0;"></span>

So the prior has pulled the inference upward slightly, from 0.65 toward the expert belief of 0.75.

:::

:::

---

🏀 **Example: Basketball with a Beta-Binomial Conjugate Prior (continue)** 


```{r}
n <- 60; x <- 39          # data
alpha <- 15; beta  <- 5   # prior

alpha_post <- alpha + x   # posterior
beta_post  <- beta + (n - x)

# Summaries
prior_mean <- alpha / (alpha + beta)
prior_mode <- (alpha - 1) / (alpha + beta - 2)

post_mean <- alpha_post / (alpha_post + beta_post)
post_mode <- (alpha_post - 1) / (alpha_post + beta_post - 2)

# 95% credible interval
ci_95 <- qbeta(c(0.025, 0.975), alpha_post, beta_post)

cat("Prior: Beta(", alpha, ", ", beta, "), Mean = ", round(prior_mean, 3),
    ", Mode = ", round(prior_mode, 3), sep = "")

cat("Posterior: Beta(", alpha_post, ", ", beta_post, "), Mean = ", round(post_mean, 3), 
    ", Mode = ", round(post_mode, 3), sep = "")

cat("95% credible interval = (", round(ci_95[1], 3), ", ", round(ci_95[2], 3), ")", sep = "")
```

---


```{r}
#| fig-align: center

curve(dbeta(x, alpha_post, beta_post), from = 0, to = 1, col = "red", lwd = 2)

curve(dbeta(x, alpha, beta), from = 0, to = 1,
      col = "darkgreen", lwd = 2, ylab = "Density", xlab = "p",
      main = "Beta-Binomial Basketball Example", add = TRUE)


abline(v = x / n, col = "blue", lty = 2, lwd = 2)
legend("topright", 
       legend = c("Prior Beta(15,5)", "Posterior Beta(54,26)", "Sample Proportion = 39/60"),
       col = c("darkgreen", "red", "blue"), lwd = c(2, 2, 2), lty = c(1, 1, 2))
```

---

## Posterior Mean Estimates: Bayesian Beta-Binomial

<span style="display: block; margin: 24px 0;"></span>

- Bayesians like to use the **posterior mean** as a point estimate. The posterior mean in this case is

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}

$$
\hat{p} = E(p\mid y) = \frac{\alpha}{\alpha + \beta} = \frac{y+a}{n+a+b}.
$$

:::

<span style="display: block; margin: 24px 0;"></span>

- This is actually a **weighted average of the MLE estimate and the prior mean**. Recall:


::: {.fragment}

$$
\hat{p}_{MLE} = \frac{y}{n}, \qquad E[p] = \frac{a}{a+b}.
$$

:::

<span style="display: block; margin: 24px 0;"></span>

::: {.fragment}

Then, the posterior mean can be expressed as

<span style="display: block; margin: 24px 0;"></span>

$$
\hat{p} = \frac{y+a}{n+a+b} = \frac{n}{n+a+b}\frac{y}{n} + \frac{a+b}{n+a+b}\frac{a}{a+b}. 
$$

<span style="display: block; margin: 48px 0;"></span>


:::

::: {.note .fragment}

This suggests **the posterior mean estimate will ALWAYS be between the MLE and the prior mean**.

:::

# Part 2 Simulation of Discrete Random Variables

## Precision Method

<span style="display: block; margin: 24px 0;"></span>

The precision method is simple and widely used for simulating discrete distributions with a small number of possible outcomes. Its main advantages are:

- Conceptually straightforward
- Easy to implement
- Works for any discrete distribution


::: {.fragment}
However, when the number of outcomes becomes very large, more efficient algorithms may be required.
:::

<span style="display: block; margin: 24px 0;"></span>

::: {.note .fragment}
Suppose $X$ is a discrete random variable that can take values

<span style="display: block; margin: 24px 0;"></span>

$$
x_1, x_2, \ldots, x_k
$$

with corresponding probabilities

<span style="display: block; margin: 24px 0;"></span>

$$
P(X=x_i)=p_i, \qquad i=1,2,\ldots,k,
$$

where

$$
p_i \ge 0, \qquad \sum_{i=1}^{k} p_i = 1.
$$

Our goal is to **generate simulated values of $X$** so that the outcomes occur with exactly these probabilities.

<span style="display: block; margin: 24px 0;"></span>

:::

---

#### Precision Method (continue)

The precision method works by **partitioning the unit interval $(0,1)$** into segments whose lengths correspond to the probabilities $p_1, p_2, \ldots, p_k$. Define the cumulative probabilities

$$
F(x_i) = \sum_{j=1}^{i} p_j, \qquad i=1,\ldots,k.
$$

::: {.fragment}

These cumulative values divide the interval $(0,1)$ into $k$ subintervals:

<span style="display: block; margin: 24px 0;"></span>

$$
(0, F(x_1)],\quad
(F(x_1), F(x_2)],\quad
\ldots,\quad
(F(x_{k-1}), F(x_k)].
$$

<span style="display: block; margin: 24px 0;"></span>

Because $F(x_k)=1$, these intervals cover the entire unit interval.

:::

::: {.fragment}

The simulation procedure is then straightforward:

:::

1. Generate a random number $U \sim \text{Uniform}(0,1)$.
2. Determine which interval contains $U$.
3. Return the corresponding value of $X$:

::: {.fragment}

$$
X =
\begin{cases}
x_1 & \text{if } 0 < U \le F(x_1) \\
x_2 & \text{if } F(x_1) < U \le F(x_2) \\
\vdots \\
x_k & \text{if } F(x_{k-1}) < U \le 1
\end{cases}
$$

::::

::: {.fragment}

This method guarantees that the probability of selecting $x_i$ is exactly $p_i$.

:::

---

#### Precision Method Algorithm

The precision method can be implemented using the following steps.

1. Specify the possible values $x_1,\ldots,x_k$ and their probabilities $p_1,\ldots,p_k$.
2. Compute the cumulative probabilities $F(x_i)$.
3. Generate a random number $U \sim \text{Uniform}(0,1)$.
4. Find the smallest $i$ such that $U \le F(x_i).$
5. Output $X = x_i$.

::: {.example .fragment}

**Example: Rolling a Loaded Die**

Suppose a die has the following probabilities:

| Outcome $x$ | 1 | 2 | 3 | 4 | 5 | 6 |
|---|---|---|---|---|---|---|
| $P(X=x)$ | 0.10 | 0.15 | 0.20 | 0.25 | 0.20 | 0.10 |

First compute the cumulative distribution:

| Outcome $x$ | 1 | 2 | 3 | 4 | 5 | 6 |
|---|---|---|---|---|---|---|
| $P(X=x)$ | 0.10 | 0.15 | 0.20 | 0.25 | 0.20 | 0.10 |
| $F(x)$ | 0.10 | 0.25 | 0.45 | 0.70 | 0.90 | 1.00 |

:::


---

::: {.example}

This corresponds to the following intervals:

:::: {.columns}
:::{.column width=31%}
| Interval | Outcome |
|:---:|:---:|
| $0 < U \le 0.10$ | 1 |
| $0.10 < U \le 0.25$ | 2 |
| $0.25 < U \le 0.45$ | 3 |
:::
:::{.column width=31%}
| Interval | Outcome |
|:---:|:---:|
| $0.45 < U \le 0.70$ | 4 |
| $0.70 < U \le 0.90$ | 5 |
| $0.90 < U \le 1.00$ | 6 |
:::
:::{.column .center width=31%}
For example:

- If $U = 0.18$, then $X=2$.
- If $U = 0.63$, then $X=4$.
- If $U = 0.92$, then $X=6$.
:::

::::

:::

::: {.fragment}


```{r}
x <- c(1,2,3,4,5,6)
p <- c(0.10,0.15,0.20,0.25,0.20,0.10)
F <- cumsum(p)   # cumulative probabilities
simulate_die <- function() {
  u <- runif(1) # uniform(0,1)
  return(x[which(u <= F)[1]]) # finds smallest i, u <= F
}

replicate(10, simulate_die()) 

# empirically approximate the target probabilities
samples <- replicate(10000, simulate_die())
table(samples) / 10000
```

:::

- In this way, the precision method provides a clear and reliable foundation for simulating any discrete random variable.


---

## Inverse CDF method

Inverse transform sampling provides a **systematic way to turn Uniform(0,1) draws into samples from any distribution with a known CDF**. The method forms the foundation for many simulation methods. It works whenever the inverse CDF is available in closed form.Otherwise, alternatives are needed when $F_X^{-1}$ is not explicit.

- Based on the **probability integral transform**: if $X$ has CDF $F_X$, then $F_X(X)\sim \text{Uniform}(0,1)$. 
  
- Inverse transform sampling **reverses** this: apply $F_X^{-1}$ to a Uniform(0,1) value to obtain a draw from $F_X$.  

<span style="display: block; margin: 24px 0;"></span>
  
::: {.note .fragment}  
**Algorithm:**
  1. Generate $U \sim \text{Uniform}(0,1). \qquad$ 2. Set $X = F_X^{-1}(U). \qquad$  3. Then $X$ has CDF $F_X$.  
:::

<span style="display: block; margin: 24px 0;"></span>

| Feature | **Precision Method** | **Inverse CDF Method** |
|-------------------|----------------------|----------------------|
| Type of distribution | Discrete | Continuous (or discrete with closed‑form inverse) |
| Core idea | Interval search on cumulative probabilities | Apply $F^{-1}(U)$ directly |
| Requires closed‑form inverse? | **No** | **Yes** |
| Implementation | `cumsum()` + find interval | Compute $F^{-1}(U)$ |
| When useful | Any discrete pmf, tables, empirical distributions | Distributions with analytic inverse CDF |
| Conceptual relationship | Discrete version of inverse CDF | General inverse transform principle |

: {.striped .hover .fragment}

---


::: {.example}

**Example: Simulating Customer Arrivals at a Service Desk**

A small service desk receives customers according to the following empirical distribution, estimated from past data:

| Number of arrivals in a 5‑minute interval | 0 | 1 | 2 | 3 | 4 |
|-------------------------------------------|---|---|---|---|---|
| Probability                               |0.05|0.25|0.40|0.20|0.10|

Management wants to **simulate one hour of operation** (12 intervals of 5 minutes) to understand typical workload patterns. Use the **discrete inverse CDF (precision) method** to generate arrivals in each interval.


:::

::: {.fragment}

```{r}
x <- 0:4 # support
p <- c(0.05, 0.25, 0.40, 0.20, 0.10) # probabilities
F <- cumsum(p) # Cumulative distribution

# Discrete inverse CDF sampler
r_arrivals <- function(n = 1) {
  u <- runif(n) # random uniform(0,1)
  x[ findInterval(u, F) + 1 ]
}
# Simulate 12 intervals (1 hour)
arrivals <- r_arrivals(12); arrivals

sum(arrivals) # Total customers in the hour
```

:::

- Function `findInterval(u,F)` finds which cumulative‑probability interval $u$ falls into.
- `x[... +1]` converts the interval index into the actual arrival count (0,1,2,3,4).



---

```{r}
#| fig-align: center
# Empirical check with a large simulation
set.seed(123)
samples <- r_arrivals(10000)
empirical <- table(samples) / 10000

barplot(rbind(p, empirical), names.arg = x,
        beside = TRUE, col = c("skyblue", "orange"),
        legend.text = c("True p(x)", "Empirical"),
        main = "Simulated vs True Arrival Distribution")
```

With 10,000 samples, the empirical distribution closely matches the target distribution—demonstrating that the discrete inverse CDF method works exactly as intended.


---

## 🦠 Case study: Epidemic SIR Modelling



SIR stands for Susceptible, Infected, and Removed. Individuals can be one of three types: 

:::: {.columns}
::: {.column .center .block .blue .fragment width=24%}
**Susceptible**<br>
if they have not yet caught the disease
:::
::: {.column .center .block .yellow .fragment width=24%}
**Infected**<br>
if they currently have the disease
:::
::: {.column .center .block .red .fragment width=40%}
**Removed**<br>
if they have had the disease and have since recovered (and are now immune) or died.
:::
::::



- Let $S(t), I(t)$ and $R(t)$ be the number of susceptible, infected and removed individuals at time t. 

- At each time step, each infected has probability $\alpha$ of
infecting each susceptible. (This assumes that each infected has equal contact with all susceptibles. This is called a mixing assumption.) 

- At the end of each time step, after having had a chance to infect people, each infected has probability of $\beta$ being removed.

::: {.fragment}

```{r}
SIRsim <- function(a, b, N, T) {
# a is infection rate, b is removal rate
# N initial susceptibles, 1 initial infected, simulation length T
# returns a matrix size (T+1)*3 with columns S, I, R respectively
S <- rep(0, T+1); I <- rep(0, T+1); R <- rep(0, T+1)
S[1] <- N; I[1] <- 1; R[1] <- 0
for (i in 1:T) {
  S[i+1] <- rbinom(1, S[i], (1 - a)^I[i])
  R[i+1] <- R[i] + rbinom(1, I[i], b)
  I[i+1] <- N + 1 - R[i+1] - S[i+1]
}
return(matrix(c(S, I, R), ncol = 3))
}
```

:::

---

Plot $S(t), I(t), R(t)$ for four separate simulations, with
$\alpha = 0.0005$ and $\beta = 0.1, 0.2, 0.3,$ and $0.4.$

```{r}
Sim1 <- SIRsim(0.0005, 0.1, 1000,100); Sim2 <- SIRsim(0.0005, 0.2, 1000,100)
Sim3 <- SIRsim(0.0005, 0.3, 1000,100); Sim4 <- SIRsim(0.0005, 0.4, 1000,100)
```


:::: {.columns}
::: {.column}
```{r}
#| code-fold: true
#| code-summary: Beta = 0.1
#| fig-height: 10

par(mfrow=c(3,1), mar=c(2,5,2,1))
plot(Sim1[,1], type='l', ylab='S(t)', cex.lab=2.5, cex.axis=2)
plot(Sim1[,2], type='l', ylab='I(t)', cex.lab=2.5, cex.axis=2)
plot(Sim1[,3], type='l', ylab='R(t)', cex.lab=2.5, cex.axis=2)
```

:::

::: {.column}

```{r}
#| code-fold: true
#| code-summary: Beta = 0.2
#| fig-height: 10

par(mfrow=c(3,1), mar=c(2,5,2,1))
plot(Sim2[,1], type='l', ylab='S(t)', cex.lab=2.5, cex.axis=2)
plot(Sim2[,2], type='l', ylab='I(t)', cex.lab=2.5, cex.axis=2)
plot(Sim2[,3], type='l', ylab='R(t)', cex.lab=2.5, cex.axis=2)
```

:::
::::

---

:::: {.columns}
::: {.column}
```{r}
#| code-fold: true
#| code-summary: Beta = 0.3
#| fig-height: 8

par(mfrow=c(3,1), mar=c(2,5,2,1))
plot(Sim3[,1], type='l', ylab='S(t)', cex.lab=2.5, cex.axis=2)
plot(Sim3[,2], type='l', ylab='I(t)', cex.lab=2.5, cex.axis=2)
plot(Sim3[,3], type='l', ylab='R(t)', cex.lab=2.5, cex.axis=2)
```

:::

::: {.column}

```{r}
#| code-fold: true
#| code-summary: Beta 0.4
#| fig-height: 8

par(mfrow=c(3,1), mar=c(2,5,2,1))
plot(Sim4[,1], type='l', ylab='S(t)', cex.lab=2.5, cex.axis=2)
plot(Sim4[,2], type='l', ylab='I(t)', cex.lab=2.5, cex.axis=2)
plot(Sim4[,3], type='l', ylab='R(t)', cex.lab=2.5, cex.axis=2)
```

:::
::::

- **As β increases**, infected individuals recover **faster**.
- Faster recovery means **less time spent infectious**, so fewer secondary infections occur. Therefore:
  - The **peak of $I(t)$** becomes **lower**.
  - The **epidemic ends sooner**.
  - The **final number of susceptibles $S(T)$** is **higher**.
  - The **epidemic size decreases**.

---

The main aim is to estimate $E[S(T)]$, average epidemic size, for different values of $\alpha$ and $\beta$ to see how it varies.

```{r}
#| code-fold: true
SIR <- function(a, b, N, T) {
# simulates SIR epidemic model from time 0 to T
# returns number of susceptibles, infected and removed at time T
S <- N
I <- 1
R <- 0
for (i in 1:T) {
S <- rbinom(1, S, (1 - a)^I)
R <- R + rbinom(1, I, b)
I <- N + 1 - S - R
}
return(c(S, I, R))
}

# set parameter values
N <- 1000
T <- 100
a <- seq(0.0001, 0.001, by = 0.0001)
b <- seq(0.1, 0.5, by = 0.05)
n.reps <- 400 # sample size for estimating E S[T]
f.name <- "SIR_grid.dat" # file to save simulation results

# estimate E S[T] for each combination of a and b
write(c("a", "b", "S_T"), file = f.name, ncolumns = 3)
for (i in 1:length(a)) {
for (j in 1:length(b)) {
S.sum <- 0
for (k in 1:n.reps) {
S.sum <- S.sum + SIR(a[i], b[j], N, T)[1]
}
write(c(a[i], b[j], S.sum/n.reps), file = f.name,
ncolumns = 3, append = TRUE)
}
}
# plot estimates in 3D
g <- read.table(f.name, header = TRUE)
library(lattice)
print(wireframe(S_T ~ a*b, data = g, scales = list(arrows = FALSE),
aspect = c(.5, 1), drape = TRUE,
xlab = "a", ylab = "b", zlab = "E S[T]"))
```



The 3D plot shows how the expected epidemic size depends jointly on infection probability and recovery probability: epidemics grow larger when α is high and β is low, and shrink when α is low and β is high.
